### 2017-04-12 doc/tyde17/main.tex
[ doc ] Beginning of a paper

%% For double-blind review submission
\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission
%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission
%\documentclass[acmlarge]{acmart}\settopmatter{}

%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format should change 'acmlarge' to
%% 'sigplan,10pt'.

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2017}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
\setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

\newcommand{\parser}[1]{\texttt{#1}}
\newcommand{\type}[1]{\texttt{#1}}

\begin{document}

%% Title information
\title{agdarsec -- Total Parser Combinators}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{Guillaume Allais}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
%  \position{Position1}
  \department{Digital Security}              %% \department is recommended
  \institution{Radboud University}            %% \institution is required
%  \streetaddress{Street1 Address1}
  \city{Nijmegen}
%  \state{State1}
%  \postcode{Post-Code1}
  \country{The Netherlands}
}
\email{gallais@cs.ru.nl}          %% \email is recommended

%% Paper note
%% The \thanks command may be used to create a "paper note" ---
%% similar to a title note or an author note, but not explicitly
%% associated with a particular element.  It will appear immediately
%% above the permission/copyright statement.
% \thanks{with paper note}                %% \thanks is optional
                                        %% can be repeated if necesary
                                        %% contents suppressed with 'anonymous'

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
\end{CCSXML}

% \ccsdesc[500]{Software and its engineering~General programming languages}
% \ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{Parser Combinators, Agda, Total Programming}  %% \keywords is optional

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

Parser combinators have made functional languages such as Haskell
shine: they are a prime example of the advantages Embedded Domain
Specific Languages provide the end user. She not only has access
to a library of powerful composable abstractions but she is also
able to rely on the existing tooling and libraries of the host
language, she gets feedback from the static analysis built in the
compiler (e.g. type and coverage checking) and can exploit from
the expressivity of the host language to write generic parsers
thanks to polymorphism and higher order functions.

\section{A Primer on Parser Combinators}

\paragraph{Parser Type}

It is naturally possible to run such a parser and try to extract
a value from a successful parse. Opinions may vary on whether a
run with leftover characters can be considered successful or not.
We decide against it. This is not crucial as both styles can be
mutually emulated by either providing an 'end of file' parser
guaranteeing that only runs with no leftovers are successful or
by extending a grammar with a dummy case consuming the rest of
the input string.

\paragraph{(Strongly-Typed) Combinators}

The most basic parser is the one that accepts any character.
It succeeds as long as the input string is non empty and
returns one result: the tail of the string together with the
character it just read.

% insert anyChar

However what makes parsers interesting is that they recognize
structure. As such, they need to reject invalid inputs. The
parser only accepting decimal digit is a bare bones example.
It can be implemented in terms of \parser{guard}, a higher
order parser checking that the value returned by its argument
abides by a predicate.

% insert guard, digit

These two definitions are only somewhat satisfactory: the
resut of the \parser{digit} parser is still stringly-typed.
Instead of using a predicate to decide whether to keep the
value, we can opt for a validation function which returns
a witness whenever the check succeeds. To define this
alternative \parser{guardM} we can once more rely on code
already part of the standard library.

In our concrete example of recognizing a digit, we want to
return an \type{Int} corresponding to it. Once more the
standard library has just the right function to use together
with \parser{guardM}.

% insert guardM, digit

\paragraph{Code Reuse: Recognizing Structures}

We have seen how we can already rely on the standard library
of the host language to seemlessly implement combinators.
We can leverage even more of the existing codebase by noticing
that the type constructor \type{Parser} is a \type{Functor},
an \type{Applicative}, a \type{Monad} and also an \type{Alternative}.

% insert alternative instance

% interesting use case? replicateM?

\paragraph{Expressivity: Higher Order Parsers and Fixpoints}

Our first example of a higher order parser was \parser{guardM}
which takes as arguments a validation function as well as another
parser and produces a parser for the type of witnesses returned
by the validation function.

% insert some, many

\section{The Issue with Haskell's Parser Types}

The ability to parse recursive grammars by simply declaring
them in a recursive manner is however dangerous: unlike type
errors which are caught by the typechecker and partial covers
in pattern matchings which are detected by the coverage checker,
termination is not guaranteed here.

The problem already shows up in the definition of \parser{some}
which will only make progress if its argument actually uses up
part of the input string. Otherwise it may loop. However this
is not the typical hurdle programmers stumble upon: demanding
a non empty list of nothing at all is after all rather silly.
The issue manifests itself naturally whenever defining a left
recursive grammar which leads us to introducing the prototypical
such example \type{Expr}:

\[
\texttt{Expr} ~::=~ \texttt{<Int>} ~|~ \texttt{<Expr>} ~\text{`+`}~ \texttt{<Expr>}
\]

If we simply reproduce this definition,

\section{Indexed Sets and Course-of-Values Recursion}

%% Acknowledgments
%\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
%  This material is based upon work supported by the
%  \grantsponsor{GS100000001}{National Science
%    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%  conclusions or recommendations expressed in this material are those
%  of the author and do not necessarily reflect the views of the
%  National Science Foundation.
%\end{acks}

%% Bibliography
%\bibliography{bibfile}

%% Appendix
%\appendix
%\section{Appendix}

\end{document}

### 2017-04-18 doc/tyde17/main.tex
[ lhs2tex ] capturing code and including it
[ agda ] capturing code and including it
Had to modify some of the source files to:
- hide implicit Set arguments using sections
- add tags to use capture the code

@@
                         %% http://ctan.org/pkg/subcaption
+\usepackage{agda}
+\usepackage{todonotes}
+\usepackage{catchfilebetweentags}
+
+\newcommand{\shgrab}[1]{\bigskip\ExecuteMetaData[parser.tex]{#1}}
+\newcommand{\hgrab}[1]{\ExecuteMetaData[parser.tex]{#1}}
+
+\newcommand{\sagrab}[2]{\bigskip\ExecuteMetaData[#1.tex]{#2}}
+\newcommand{\agrab}[2]{\ExecuteMetaData[#1.tex]{#2}}
+
+\input{commands.tex}
+\input{lhs2tex-prelude.tex}
 
@@
 
+\shgrab{parser}
+
 It is naturally possible to run such a parser and try to extract
@@
 
+\shgrab{parse}
+
 \paragraph{(Strongly-Typed) Combinators}
@@
 
-% insert anyChar
+\shgrab{anychar}
 
@@
 
-% insert guard, digit
+\shgrab{guard}
+\hgrab{digit}
 
@@
 
-% insert guardM, digit
+\shgrab{guard2}
+\hgrab{digit2}
 
@@
 that the type constructor \type{Parser} is a \type{Functor},
-an \type{Applicative}, a \type{Monad} and also an \type{Alternative}.
+an \type{Applicative}~\todo{cite}, a \type{Monad} and also an \type{Alternative}.
 
@@
 
-% insert some, many
+\shgrab{some}
+\hgrab{many}
 
@@
 recursive grammar which leads us to introducing the prototypical
-such example \type{Expr}:
+such example: \type{Expr}, a minimal language of arithmetic
+expressions.
 
@@
 
-If we simply reproduce this definition,
+The intuitive solution is to simply reproduce this definition by
+introducing an inductive type for \type{Expr} and then defining
+the parser as an alternative between a literal on one hand and a
+sub-expression, the character '+', and another sub-expression on
+the other.
+
+\shgrab{Expr}
+\hgrab{expr}
+
+However this leads to an infinite loop. Indeed, the second
+alternative performs a recursive call to \parser{expr} even
+though it hasn't consumed any character from the input string.
+
+The typical solution to this problem is to introduce two 'tiers'
+of expressions: the \parser{base} ones which can only be whole
+expressions if we consume an opening parenthesis first and the
+\parser{expr'} ones which are left-associated chains of \parser{base}
+expressions connected by '+'.
+
+\shgrab{base}
+\hgrab{expr2}
+
+This approach can be generalised when defining more complex
+languages by having even more tiers, one for each
+\emph{fixity level}. An extended language of arithmetic
+expressions would for instance distinguish the level at
+which addition and substraction live from the one at which
+multiplication and division do.
+
+Our issue with this solution is twofold. Firt, although we did
+eventually managed to build a parser that worked as expected,
+the compiler was unable to warn us and guide us towards this
+correct solution. Additionnally, the blatant partiality of
+some of these definitions means that these combinators and
+these types are wholly unsuitable in a total. We could, of
+course use an escape hatch and implement our parsers in Haskell
+but that would both be unsafe and mean we would not be able
+to run them at typechecking time which we may want to do if
+we embed checked examples in our software's documentation,
+or use constant values (e.g. filepaths).
 
 \section{Indexed Sets and Course-of-Values Recursion}
+
+Our implementation of Total Parser Combinators is in Agda,
+a total dependently typed programming language and it will
+rely heavily on indexed sets. But the indices will not be
+playing any interesting role apart from enforcing totality.
+As a consequence, we introduce combinators to build indexed
+sets without having to mention the index explicitly. This
+ought to make the types more readable by focusing on the
+important components and hiding away the artefacts of the
+encoding.
+
+The first kind of combinator corresponds to operations on
+sets which are lifted to indexed sets by silently propagating
+the index. It may be a bit surprising to see infix notations
+(in Agda underscores correspond to positions in which arguments
+are to be inserted) for functions taking three arguments but
+they are only meant to be partially applied.
+
+\sagrab{Relation/Unary/Indexed}{arrow}
+\agrab{Relation/Unary/Indexed}{product}
+
+The second kind turns an indexed set into a set by universally
+quantifying over the index. Here we use curly braces so that
+the index we use is an \emph{implicit} argument we will never
+have to write.
+
+\sagrab{Relation/Unary/Indexed}{forall}
+
+Finally, the last kind of combinator\todo{cite}
+Instead of defining it as a
+
+\sagrab{Induction/Nat/Strong}{box}
+
+\sagrab{Induction/Nat/Strong}{extract}
+\agrab{Induction/Nat/Strong}{duplicate}
+
+\sagrab{Induction/Nat/Strong}{map}
+\agrab{Induction/Nat/Strong}{app}
+
+\agrab{Induction/Nat/Strong}{fix}
+
+\paragraph{Remark} The type of \AF{fix} corresponds exactly to
+strong induction for the natural numbers. A similar \AR{□} type
+constructor can be defined for any induction principle relying
+on an accessibility predicate. Which means that a library's
+types can be cleaned up by using these combinators in any
+situation where one had to give up structural induction for a
+more powerful alternative.
+
+\section{Parsing, Totally}
 

### 2017-04-19 doc/tyde17/main.tex
[ doc ] more about box

@@
 
-\newcommand{\shgrab}[1]{\bigskip\ExecuteMetaData[parser.tex]{#1}}
+\newcommand{\shgrab}[1]{\medskip\ExecuteMetaData[parser.tex]{#1}}
 \newcommand{\hgrab}[1]{\ExecuteMetaData[parser.tex]{#1}}
 
-\newcommand{\sagrab}[2]{\bigskip\ExecuteMetaData[#1.tex]{#2}}
+\newcommand{\sagrab}[2]{\medskip\ExecuteMetaData[#1.tex]{#2}}
 \newcommand{\agrab}[2]{\ExecuteMetaData[#1.tex]{#2}}
+\newcommand{\BOX}{\AR{□}}
 
@@
 
-Finally, the last kind of combinator\todo{cite}
-Instead of defining it as a
+We can already see the benefits of using these aliases. For instance
+the fairly compact expression \AF{[} (\AB{P} \AF{⊗} \AB{Q}) \AF{⟶} \AB{R} \AF{]}
+corresponds to the more verbose type
+\AS{∀} \{\AB{n}\} \AS{→} (\AB{P} \AB{n} \AF{×} \AB{Q} \AB{n}) \AS{→} \AB{R} \AB{n}.
+
+Last but not least, the \BOX{} type constructor takes a \AD{ℕ}-indexed
+set and produces the one corresponding to the allowed recursive
+calls in a function defined by course-of-values recursion: \BOX{}
+\AB{A} \AB{n} associates to each \AB{m} strictly smaller than \AB{n}
+a value of type \AB{A} \AB{m}.
 
@@
 
-\sagrab{Induction/Nat/Strong}{extract}
-\agrab{Induction/Nat/Strong}{duplicate}
+\paragraph{Remark: Record Wrapper} Instead of defining \BOX{} as a
+function like the other combinators, we wrap the corresponding
+function space in a record type. This prevents normalisation from
+unfolding the combinator too eagerly and makes types more readable
+during interactive development.
+
+\paragraph{Remark: Irrelevance} The argument stating that \AB{m}
+is strictly smaller than \AB{n} is preceded by a dot. In Agda,
+it means that this value is irrelevant and can be erased by the
+compiler. In Coq, we would define the relation \AF{\_<\_} in
+\type{Prop} to achieve the same.
+
+\medskip{}
+
+This \BOX{} type combinator has some useful value combinators.
+The first thing we can notice is the fact that \textbf{\BOX{} is a
+functor}; that is to say that given a natural transformation from
+\AB{A} to \AB{B}, we can define a natural transformation from
+\BOX{} \AB{A} to \BOX{} \AB{B}.
 
 \sagrab{Induction/Nat/Strong}{map}
+
+\paragraph{Remark: Copatterns} The definition of \AF{map} uses
+the \BOX{} field named \ARF{call} on the \emph{left hand side}.
+This is a copattern\todo{cite}, meaning that we explain how
+the definition is \emph{observed} (via \ARF{call}) rather than
+\emph{constructed} (via \AIC{mkBox}).
+
+\medskip{}
+
+Because \AF{\_<\_} is defined in terms of \AD{\_≤\_}, \AF{≤-refl}
+which is the proof that \AD{\_≤\_} is reflexive is also a proof
+that any \AB{n} is strictly smaller than \AN{1} \AF{+} \AB{n}.
+We can use this fact to write the following \AF{extract} function:
+
+\sagrab{Induction/Nat/Strong}{extract}
+
+\paragraph{Remark: Counit} The careful reader will have noticed
+that this is not quite the extract we would expect from a comonad:
+for a counit, we would need a natural transformation between
+\BOX{} \AB{A} and \AB{A} i.e. a function of type
+\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}. We will not be able
+to define such a function: \BOX{} \AB{A} \AN{0} is isomorphic to
+the unit type so we would have to generate an \AB{A} \AN{0} out
+of thin air.
+
+\medskip
+
+Even though we cannot have a counit, we are still able to define
+a comultiplication thanks to the fact that \AF{\_<\_} is transitive.
+
+\sagrab{Induction/Nat/Strong}{duplicate}
+
 \agrab{Induction/Nat/Strong}{app}
+\sagrab{Induction/Nat/Strong}{fix}
 
-\agrab{Induction/Nat/Strong}{fix}
-
-\paragraph{Remark} The type of \AF{fix} corresponds exactly to
-strong induction for the natural numbers. A similar \AR{□} type
-constructor can be defined for any induction principle relying
-on an accessibility predicate. Which means that a library's
-types can be cleaned up by using these combinators in any
-situation where one had to give up structural induction for a
-more powerful alternative.
+\paragraph{Remark: Generalisation} The type of \AF{fix} corresponds
+exactly to strong induction for the natural numbers. A similar \BOX{}
+type constructor can be defined for any induction principle relying
+on an accessibility predicate. Which means that a library's types
+can be cleaned up by using these combinators in any situation where
+one had to give up structural induction for a more powerful alternative.
 

### 2017-10-02 doc/paper/main.tex
[ doc ] start of refs

--- 2017-04-19 doc/tyde17/main.tex
@@
 Last but not least, the \BOX{} type constructor takes a \AD{ℕ}-indexed
-set and produces the one corresponding to the allowed recursive
-calls in a function defined by course-of-values recursion: \BOX{}
-\AB{A} \AB{n} associates to each \AB{m} strictly smaller than \AB{n}
-a value of type \AB{A} \AB{m}.
+set and produces the set of valid recursive calls for a function defined
+by course-of-values recursion: \BOX{} \AB{A} \AB{n} associates to each
+\AB{m} strictly smaller than \AB{n} a value of type \AB{A} \AB{m}.
+This construct, analogous to the later modality showing up in Guarded
+Type Theory~\cite{vezzosi2015guarded}, empowers the user to give precise
+types in a total language to programs commonly written in partial ones
+(see e.g. the definition of \textit{fix} below).
 
@@
 
+\paragraph{Remark: Identifiers in Agda} In Agda, any space-free
+string which is not a reserved keyword is a valid identifier. As
+a consequence we can pick suggestive names such as \AB{m<n} for
+a proof that $m < n$.
+
 \agrab{Induction/Nat/Strong}{app}
@@
 %% Bibliography
-%\bibliography{bibfile}
+\bibliography{main.bib}
 

### 2017-10-13 doc/paper/main.tex
[ more ] start of parsing, totally

@@
 Parser combinators have made functional languages such as Haskell
-shine: they are a prime example of the advantages Embedded Domain
+shine. They are a prime example of the advantages Embedded Domain
 Specific Languages provide the end user. She not only has access
-to a library of powerful composable abstractions but she is also
-able to rely on the existing tooling and libraries of the host
-language, she gets feedback from the static analysis built in the
-compiler (e.g. type and coverage checking) and can exploit from
-the expressivity of the host language to write generic parsers
-thanks to polymorphism and higher order functions.
+to a library of powerful and composable abstractions but she is also
+able to rely on the host language's existing tooling and libraries.
+She also gets feedback from the static analyses built in the compiler
+(e.g. type and coverage checking) and can exploit the expressivity
+of the host language to write generic parsers thanks to polymorphism
+and higher order functions.
+
+However she only gets the guarantees the host language is willing
+to give. In non-total programming languages such as Haskell this
+means she will not be prevented from writing parsers which will
+unexpectedly fail on some (or even all!) inputs. Handling a
+left-recursive grammar is perhaps the most iconic pitfall leading
+beginners to their doom: a parser never making any progress.
+
+We start with a primer on parser combinators and follow up with the
+definition of a broken parser which is silently accepted by Haskell.
+We then move on to Agda~\todo{cite} and introduce combinators to
+define functions by well-founded recursion. This allows us to define
+a more informative notion of parser and give more precise types to
+the combinators commonly used. We then demonstrate that broken parsers
+such as the one presented earlier are rejected whilst typical example
+can be ported with minimal modifications.
 
@@
 
-\paragraph{Parser Type}
+\paragraph{Parser Type} Let us start by reminding ourselves what
+a parser is. Although we will eventually move to a more generic
+type, Fritz Ruehr's rhyme gives us the \emph{essence} of parsers:
+
+\begin{verse}
+  A Parser for Things\\
+  is a function from Strings\\
+  to Lists of Pairs\\
+  of Things and Strings!
+\end{verse}
+
+This stanza directly translates to the following Haskell type (we
+use a \textbf{newtype} wrapper to have cleaner error messages):
 
@@
 run with leftover characters can be considered successful or not.
-We decide against it. This is not crucial as both styles can be
+We decide against it.\todo{ambiguity} This is not crucial as both styles can be
 mutually emulated by either providing an 'end of file' parser
@@
 % insert alternative instance
-
 % interesting use case? replicateM?
@@
 
+The two parsers \AB{some} and \AB{many} turn a parser for elements
+into ones for non-empty and potentially empty lists of such elements
+respectively. They concisely showcase the power of mutual recursion,
+higher-order functions and the \type{Functor}, \type{Applicative},
+and \type{Alternative} structure.
+
 \shgrab{some}
@@
 
-\section{The Issue with Haskell's Parser Types}
+\section{The Issue with Haskell's Parser Types}\label{sec:haskellproblem}
 
@@
 
-\paragraph{Remark: Record Wrapper} Instead of defining \BOX{} as a
+\remark{Record Wrapper} Instead of defining \BOX{} as a
 function like the other combinators, we wrap the corresponding
@@
 
-\paragraph{Remark: Irrelevance} The argument stating that \AB{m}
+\remark{Irrelevance} The argument stating that \AB{m}
 is strictly smaller than \AB{n} is preceded by a dot. In Agda,
@@
 
-\paragraph{Remark: Copatterns} The definition of \AF{map} uses
+\remark{Copatterns} The definition of \AF{map} uses
 the \BOX{} field named \ARF{call} on the \emph{left hand side}.
-This is a copattern\todo{cite}, meaning that we explain how
-the definition is \emph{observed} (via \ARF{call}) rather than
-\emph{constructed} (via \AIC{mkBox}).
+This is a copattern~\cite{abel2013copatterns}, meaning that we
+explain how the definition is \emph{observed} (via \ARF{call})
+rather than \emph{constructed} (via \AIC{mkBox}).
 
@@
 
-\paragraph{Remark: Counit} The careful reader will have noticed
+\remark{Counit} The careful reader will have noticed
 that this is not quite the extract we would expect from a comonad:
@@
 
-\medskip
+\medskip{}
 
@@
 
-\paragraph{Remark: Identifiers in Agda} In Agda, any space-free
-string which is not a reserved keyword is a valid identifier. As
-a consequence we can pick suggestive names such as \AB{m<n} for
-a proof that $m < n$.
+\remark{Identifiers in Agda} Any space-free string which
+is not a reserved keyword is a valid identifier. As a consequence
+we can pick suggestive names such as \AB{m<n} for a proof that $m < n$.
+
+\medskip{}
+
+Exploring further the structure of the functor \BOX{}, we can observe
+that just like it is not quite a comonad, it is not quite an applicative
+functor. Indeed we can only define \AB{pure}, a natural transformation
+of type \AF{[} \AB{A} \AF{⟶} \BOX{} \AB{A} \AF{]}, for the types \AB{A}
+that are downwards closed. Providing the user with \AB{app} is however
+possible:
 
 \agrab{Induction/Nat/Strong}{app}
+
+Finally, we can reach what will serve as the backbone of our parser
+definitions: a safe, total fixpoint combinator. It differs from the
+traditional $Y$ combinator in that all the recursive calls have to be
+guarded.
+
 \sagrab{Induction/Nat/Strong}{fix}
 
-\paragraph{Remark: Generalisation} The type of \AF{fix} corresponds
+\remark{Generalisation} If we were to unfold all the type-level
+combinators and record wrappers, the type of \AF{fix} would correspond
 exactly to strong induction for the natural numbers. A similar \BOX{}
@@
 \section{Parsing, Totally}
+
+As already highlighted in Section~\ref{sec:haskellproblem}, \AB{some}
+and \AB{many} can yield diverging computations if the parser they are
+given as an argument succeeds on the empty string. To avoid any such
+issue, we adopt a radical solution: for a parser's run to be considered
+successful, it must have consumed some of its input.
+
+\agrab{Text/Parser/Success}{success}
+
+\agrab{Text/Parser/Combinators}{parser}
 

### 2017-10-17 doc/paper/main.tex
[ doc ] some changes I had forgotten to push

@@
 \usepackage{catchfilebetweentags}
+
+\DeclareUnicodeCharacter{8759}{:\,:}
 
@@
 
-\remark{Irrelevance} The argument stating that \AB{m}
-is strictly smaller than \AB{n} is preceded by a dot. In Agda,
-it means that this value is irrelevant and can be erased by the
-compiler. In Coq, we would define the relation \AF{\_<\_} in
-\type{Prop} to achieve the same.
+\remark{Irrelevance} The argument stating that \AB{m} is strictly
+smaller than \AB{n} is preceded by a dot. In Agda, it means that
+this value is irrelevant and can be erased by the compiler.
+In Coq, we would define the relation \AF{\_<\_} in \type{Prop} to
+achieve the same.
 
@@
 
+This can be made formal with the \AR{Success} record type: a \AF{Success}
+of type \AB{A} and size {n} is a value of type \AB{A} together with the
+leftovers of the input string of size strictly smaller than \AB{n}.
+
 \agrab{Text/Parser/Success}{success}
 
+\remark{Implicit Field} Just like the arguments to a function can be
+implicit, so can a record's fields. The user can them leave them out
+when building a value and they will be filled in by unification.
+
+\medskip{}
+
+Coming back to Fritz Ruehr's rhyme, we can define our own \AR{Parser}
+type: a parser for things up to size \AB{n} is a function from strings
+of length \AB{m} less than \AB{n} to lists of \AR{Success}es of size \AB{m}.
+
 \agrab{Text/Parser/Combinators}{parser}
+
+\agrab{Text/Parser/Combinators}{anyChar}
+
+\agrab{Text/Parser/Combinators}{guardM}
 

### 2017-10-26 doc/paper/main.tex
[ doc ] more about the total parser combinators

@@
 \DeclareUnicodeCharacter{8759}{:\,:}
+\DeclareUnicodeCharacter{954}{$\kappa$}
 
@@
 
-The first kind of combinator corresponds to operations on
+The first kind of combinators corresponds to operations on
 sets which are lifted to indexed sets by silently propagating
-the index. It may be a bit surprising to see infix notations
-(in Agda underscores correspond to positions in which arguments
-are to be inserted) for functions taking three arguments but
-they are only meant to be partially applied.
+the index. We only show the ones we will use in this paper:
+the pointwise arrow and product types and the constant function.
+It may be a bit surprising to see infix notations (in Agda
+underscores correspond to positions in which arguments are to
+be inserted) for functions taking three arguments but they are
+only meant to be partially applied.
 
@@
 \agrab{Relation/Unary/Indexed}{product}
+\agrab{Relation/Unary/Indexed}{constant}
 
@@
 We can already see the benefits of using these aliases. For instance
-the fairly compact expression \AF{[} (\AB{P} \AF{⊗} \AB{Q}) \AF{⟶} \AB{R} \AF{]}
+the fairly compact expression \AF{[} (\AF{κ} \AB{P} \AF{⊗} \AB{Q}) \AF{⟶} \AB{R} \AF{]}
 corresponds to the more verbose type
-\AS{∀} \{\AB{n}\} \AS{→} (\AB{P} \AB{n} \AF{×} \AB{Q} \AB{n}) \AS{→} \AB{R} \AB{n}.
+\AS{∀} \{\AB{n}\} \AS{→} (\AB{P} \AF{×} \AB{Q} \AB{n}) \AS{→} \AB{R} \AB{n}.
 
@@
 the unit type so we would have to generate an \AB{A} \AN{0} out
-of thin air.
+of thin air. The types \AB{A} for which \BOX{} has a counit are
+interesting in their own right: they are inhabited at every single
+index as demonstrated by \AF{fix} later on.
 
@@
 \remark{Implicit Field} Just like the arguments to a function can be
-implicit, so can a record's fields. The user can them leave them out
+implicit, so can a record's fields. The user can then leave them out
 when building a value and they will be filled in by unification.
@@
 
+Now that we have a precise definition of \AR{Parser}s, we can start
+building our library of combinators. Our first example \AF{anyChar}
+can be defined by copattern-matching and then case analysis on the
+input string: if it is empty then the list of \AR{Success}es is also
+empty, otherwise it contains exactly one element which corresponds to
+the head of the input string and its tail as leftovers.
+
 \agrab{Text/Parser/Combinators}{anyChar}
 
+Unsurprisingly \AF{guardM} is still a valid higher-order combinator:
+filtering out results which do not agree with a predicate is absolutely
+compatible with the consumption constraint we have drawn. To implement
+\AF{guardM} we can once more reuse existing library functions such
+\AF{gfilter} which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
+provided a predicate \mbox{\AB{A} \AS{→} \AD{Maybe} \AB{B}}.
+
 \agrab{Text/Parser/Combinators}{guardM}
+
+Demonstrating that \AR{Parser} is a functor goes along the same
+lines: using \AD{List}'s and \AR{Success}'s \AF{map}s. Similarly,
+we can prove that it is an \AF{Alternative}: failing corresponds
+to returning the empty list no matter what whilst disjunction is
+implemented using concatenation.
+
+\agrab{Text/Parser/Combinators}{fmap}\vspace*{-2em}
+\agrab{Text/Parser/Combinators}{fail}\vspace*{-2em}
+\agrab{Text/Parser/Combinators}{disjunction}
+
+So far the types we have ascribed to our combinators are exactly the
+same as the ones one would find in any other parsec library. In none
+of the previous combinators do we run a second parser on the leftovers
+of a first one. All we do is either manipulate or combine the results
+of one or more parsers run in parallel, potentially discarding some of
+these results on the way.
+
+However when we run a parser \emph{after} some of the input has already
+been consumed, we could safely perform a \emph{guarded} call. This being
+made explicit would be useful when using \AF{fix} to define a parser for
+a recursive grammar. Luckily \AR{Parser} is, by definition, a
+downwards-closed type. This means that we may use very precise types marking
+all the guarded positions with \BOX{}; if the user doesn't need that extra
+power she can very easily bypass the \BOX{} annotations by using \AF{box}:
+
+\agrab{Text/Parser/Combinators}{box}
+
+The most basic example we can give of such an annotation is probably the
+definition of a conjunction combinator \AF{\_<\&>\_} taking two parsers,
+running sequentially and returning a pair of their results. The second
+parser is given the type \mbox{\BOX{} \AR{Parser} \AB{B}} instead of
+\mbox{\AR{Parser} \AB{B}} which we would expect to find in other parsec libraries.
+
+\agrab{Text/Parser/Combinators}{conjunction}
+
+We can immediately use all of these newly-defined combinator to give
+a safe, total definition of \AF{some} which takes a parser for \AB{A}
+and returns a parser for non-empty lists of \AB{A}s. It is defined as
+a fixpoint and proceeds as follows: it either combines a head and a
+non-empty tail using \AF{$\_::^{+}\_$} or returns a singleton list.
+
+\agrab{Text/Parser/Combinators}{badsome}
+
+\remark{Inefficiency} Unfortunately this definition is inefficient.
+Indeed, in the base case \mbox{\AF{some} \AB{p}} is going to run the parser
+\AB{p} twice: once in the first branch before realising that \AB{rec}
+fails and once again in the second branch. Compare this to the Haskell
+version (after inlining \AB{many}):
+
+\shgrab{some2}
+
+This inefficiency can be fixed by using a notion of conjunction
+\AF{\_<\&?>\_} which allows the second parser to fail whilst still
+producing a successful run. The parser \mbox{\AF{some} \AB{p}} can then be
+defined as the conjunction of \AB{p} and a potentially failing
+recursive call. The function \AF{cons} (whose definition we omit here)
+either uses \AF{$\_::^{+}\_$} to put the head and the non-empty
+tail together or returns a singleton list if the recursive call
+failed.
+
+\agrab{Text/Parser/Combinators}{conjunction2}
+\agrab{Text/Parser/Combinators}{goodsome}
+
+\remark{Non-compositionality} The higher-order parser \AF{some}
+expects a \emph{fully} defined parser as an argument. This makes
+it impossible to use it as one of the building blocks of a larger,
+recursive parser. Ideally we would rather have a combinator of
+type \mbox{\AF{[} \AF{Parser} \AB{A} \AF{⟶} \AF{Parser} (\AF{List$^+$} \AB{A}) \AF{]}}.
+This will be adressed in the next subsection.
+\medskip{}
+
+As the previous code snippet shows, \AF{\_<\&?>\_} is defined in
+terms of a more fundamental notion \AF{\_\&?>>=\_} which is a
+combinator analogous to a monad's \textit{bind}. On top of running
+two parsers sequentially (with the second one being chosen based on
+the result of running the first), it allows the second one to fail
+and returns both results.
+
+\agrab{Text/Parser/Combinators}{mbind}
+
+This crucial definition makes it possible to port a lot of the Haskell
+definitions where one would use a parser which does not use any of its
+input. And this is possible without incurring any additional cost as
+the optimised version of \AF{some} showed.
+
+\subsection{Left and Right Chains}
+
+The pattern used in the solution presented in section~\ref{sec:haskellproblem}
+can be abstracted with the notion of an (heterogeneous) left chain which takes
+a parser for a seed, one for a constructor, and one for and argument. The crucial
+thing is to make sure \emph{not} to use the parser one is currently defining as
+the seed.
+
+\shgrab{hchainl}
+
+We naturally want to include a safe variant of this combinator in our
+library. However this definition relies on the ability to simply use
+\textit{return} in case it's not possible to parse an constructor
+additional constructor and argument and that is something we simply
+don't have access to.
 

### 2017-10-26 doc/paper/main.tex
[ doc ] adding an abstract

@@
 \begin{abstract}
+Parser combinator libraries represent parsers as functions and,
+using higher-order functions, define a DSL of combinators allowing
+users to quickly put together programs capable of handling complex
+grammars. When moving to total functional languages such as Agda,
+these programs cannot be directly ported: there is nothing in the
+original definitions guaranteeing termination.
+
+In this paper, we will introduce a 'guarded' modal operator acting
+on types and show how it allows us to give more precise types to
+existing combinators thus guaranteeing totality. The resulting
+library is available online together with various usage examples
+at \url{https://github.com/gallais/agdarsec}.
 \end{abstract}

### 2017-10-27 doc/paper/main.tex
[ cosmetic + references ]

--- 2017-10-26 doc/paper/main.tex
@@
 
-\newcommand{\parser}[1]{\texttt{#1}}
+\newcommand{\parser}[1]{\textit{#1}}
 \newcommand{\type}[1]{\texttt{#1}}
@@
 \begin{abstract}
+Parser combinator libraries represent parsers as functions and,
+using higher-order functions, define a DSL of combinators allowing
+users to quickly put together programs capable of handling complex
+recursive grammars. When moving to total functional languages such
+as Agda, these programs cannot be directly ported: there is nothing
+in the original definitions guaranteeing termination.
+
+In this paper, we will introduce a `guarded' modal operator acting
+on types and show how it allows us to give more precise types to
+existing combinators thus guaranteeing totality. The resulting
+library is available online together with various usage examples
+at \url{https://github.com/gallais/agdarsec}.
 \end{abstract}
@@
 definition of a broken parser which is silently accepted by Haskell.
-We then move on to Agda~\todo{cite} and introduce combinators to
+We then move on to Agda~\cite{norell2009dependently} and introduce combinators to
 define functions by well-founded recursion. This allows us to define
@@
 
-\paragraph{Parser Type} Let us start by reminding ourselves what
-a parser is. Although we will eventually move to a more generic
-type, Fritz Ruehr's rhyme gives us the \emph{essence} of parsers:
+\subsection{Parser Type}
+
+Let us start by reminding ourselves what a parser is. Although we
+will eventually move to a more generic type, Fritz Ruehr's rhyme
+gives us the \emph{essence} of parsers:
 
@@
 run with leftover characters can be considered successful or not.
-We decide against it.\todo{ambiguity} This is not crucial as both styles can be
+We decide against it. This is not crucial as both styles can be
 mutually emulated by either providing an 'end of file' parser
@@
 
-\paragraph{(Strongly-Typed) Combinators}
+\subsection{(Strongly-Typed) Combinators}
 
@@
 order parser checking that the value returned by its argument
-abides by a predicate.
+abides by a predicate which can easily be implemented using
+functions from the standard library.
 
@@
 These two definitions are only somewhat satisfactory: the
-resut of the \parser{digit} parser is still stringly-typed.
+resut of the \parser{digit} parser is still \emph{stringly-typed}.
 Instead of using a predicate to decide whether to keep the
-value, we can opt for a validation function which returns
+value, we can opt for a validation function of type
+\mbox{\textit{a → Maybe b}} which returns
 a witness whenever the check succeeds. To define this
@@
 
-\paragraph{Code Reuse: Recognizing Structures}
+\subsection{Code Reuse: Recognizing Structures}
 
@@
 that the type constructor \type{Parser} is a \type{Functor},
-an \type{Applicative}~\todo{cite}, a \type{Monad} and also an \type{Alternative}.
+an \type{Applicative}~\cite{mcbride2008applicative}, a \type{Monad} and also an \type{Alternative}.
 
@@
 
-\paragraph{Expressivity: Higher Order Parsers and Fixpoints}
+\subsection{Expressivity: Higher Order Parsers and Fixpoints}
 
@@
 \hgrab{many}
+
+\remark{Non-Commutativity} The disjunction combinator is
+non-commutative. As such the definitions of \textit{some} and
+\textit{many} will try to produce the longest list as possible
+as opposed to a flipped version of \text{many} which would start
+by returning the empty list and slowly offer longer and longer
+matches.
 
@@
 languages by having even more tiers, one for each
-\emph{fixity level}. An extended language of arithmetic
-expressions would for instance distinguish the level at
-which addition and substraction live from the one at which
-multiplication and division do.
+\emph{fixity level}, see for instance~\ref{example:expr}.
+An extended language of arithmetic expressions would for
+instance distinguish the level at which addition and
+substraction live from the one at which multiplication and
+division do.
 
@@
 some of these definitions means that these combinators and
-these types are wholly unsuitable in a total. We could, of
+these types are wholly unsuitable in a total setting. We could, of
 course use an escape hatch and implement our parsers in Haskell
@@
 the pointwise arrow and product types and the constant function.
-It may be a bit surprising to see infix notations (in Agda
-underscores correspond to positions in which arguments are to
-be inserted) for functions taking three arguments but they are
-only meant to be partially applied.
 
@@
 
-The second kind turns an indexed set into a set by universally
-quantifying over the index. Here we use curly braces so that
-the index we use is an \emph{implicit} argument we will never
-have to write.
+\remark{Mixfix Operators} In Agda underscores correspond to
+positions in which arguments are to be inserted. It may be a
+bit surprising to see infix notations for functions taking
+three arguments but they are only meant to be partially applied.
+\medskip{}
+
+The second kind of combinator turns an indexed set into a set
+by universally quantifying over the index.
 
 \sagrab{Relation/Unary/Indexed}{forall}
+
+\remark{Implicit Arguments} We use curly braces so that the
+index we use is an \emph{implicit} argument we will never
+have to write: Agda will fill it in for us by unification.
+\medskip{}
 
@@
 \remark{Counit} The careful reader will have noticed
-that this is not quite the extract we would expect from a comonad:
-for a counit, we would need a natural transformation between
-\BOX{} \AB{A} and \AB{A} i.e. a function of type
-\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}. We will not be able
+that this is not quite the \textit{extract} we would expect from a
+comonad: for a counit, we would need a natural transformation
+between \BOX{} \AB{A} and \AB{A} i.e. a function of type
+\mbox{\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}}. We will not be able
 to define such a function: \BOX{} \AB{A} \AN{0} is isomorphic to
@@
 is not a reserved keyword is a valid identifier. As a consequence
-we can pick suggestive names such as \AB{m<n} for a proof that $m < n$.
+we can pick suggestive names such as \AB{m<n} for a proof that $m < n$
+(notice the extra spaces around the infix operator $(<)$).
 
@@
 functor. Indeed we can only define \AB{pure}, a natural transformation
-of type \AF{[} \AB{A} \AF{⟶} \BOX{} \AB{A} \AF{]}, for the types \AB{A}
+of type \mbox{\AF{[} \AB{A} \AF{⟶} \AR{\BOX{}} \AB{A} \AF{]}}, for the types \AB{A}
 that are downwards closed. Providing the user with \AB{app} is however
@@
 
-\remark{Generalisation} If we were to unfold all the type-level
-combinators and record wrappers, the type of \AF{fix} would correspond
-exactly to strong induction for the natural numbers. A similar \BOX{}
+If we were to unfold all the type-level combinators and record wrappers,
+the type of \AF{fix} would correspond exactly to strong induction for
+the natural numbers. Hence its implementation also follows the one of strong
+induction: it is a combination of a call to \AF{extract} and an auxiliary
+definition \AF{fix\BOX{}} of type \mbox{\AF{[} \AR{\BOX{}} \AB{A} \AF{⟶}
+\AB{A} \AF{]} \AS{→} \AF{[} \AR{\BOX} \AB{A} \AF{]}}.
+
+\remark{Generalisation}  A similar \BOX{}
 type constructor can be defined for any induction principle relying
@@
 compatible with the consumption constraint we have drawn. To implement
-\AF{guardM} we can once more reuse existing library functions such
+\AF{guardM} we can once more reuse existing library functions such as
 \AF{gfilter} which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
@@
 
-So far the types we have ascribed to our combinators are exactly the
+So far the types we have ascribed to our combinators are, if we ignore
+the $\mathbb{N}$ indices, exactly the
 same as the ones one would find in any other parsec library. In none
@@
 definition of a conjunction combinator \AF{\_<\&>\_} taking two parsers,
-running sequentially and returning a pair of their results. The second
+running them sequentially and returning a pair of their results. The second
 parser is given the type \mbox{\BOX{} \AR{Parser} \AB{B}} instead of
@@
 
-We can immediately use all of these newly-defined combinator to give
+We can immediately use all of these newly-defined combinators to give
 a safe, total definition of \AF{some} which takes a parser for \AB{A}
@@
 \AB{p} twice: once in the first branch before realising that \AB{rec}
-fails and once again in the second branch. Compare this to the Haskell
-version (after inlining \AB{many}):
+fails and once again in the second branch. Compare this definition to
+the Haskell version (after inlining \AB{many}) where \AB{p} is run once
+and then its result is either combined with a list obtained by recursion
+or returned as a singleton:
 
@@
 two parsers sequentially (with the second one being chosen based on
-the result of running the first), it allows the second one to fail
+the result obtained by running the first), it allows the second one to fail
 and returns both results.
@@
 
-The pattern used in the solution presented in section~\ref{sec:haskellproblem}
+The pattern used in the solution presented in Section~\ref{sec:haskellproblem}
 can be abstracted with the notion of an (heterogeneous) left chain which takes
@@
 don't have access to.
+
+\subsection{Examples}
+
+\todo{expr}\label{example:expr}
 

### 2017-10-27 doc/paper/main.tex
[ more ] schainl

@@
 
-\remark{Non-compositionality} The higher-order parser \AF{some}
+\remark{Non-compositionality}\label{parser:some} The higher-order parser \AF{some}
 expects a \emph{fully} defined parser as an argument. This makes
@@
 library. However this definition relies on the ability to simply use
-\textit{return} in case it's not possible to parse an constructor
-additional constructor and argument and that is something we simply
-don't have access to.
+\textit{return} in case it's not possible to parse an additional
+constructor and argument and that is something we simply don't have
+access to. This forces us to find the essence of \parser{rest}, the
+auxiliary definition used in \parser{hchainl}: its first argument is
+not just a value, it is a \AR{Success} upon which it builds until it
+can't anymore and simply returns.
+
+We define \parser{schainl} as the fixpoint of either \parser{rest}
+(whose definition we omit here) or the \AR{Success} that was passed
+as an input (the function \AF{\_::$^{r}$} appends a single element
+at the end of a list).
+
+\agrab{Text/Parser/Combinators}{schainl}
+
+The type of \AF{rest} may look intimidating but it has this shape because
+we build a list of \AR{Success}es at the \emph{same} index as the input
+\AR{Success} and \AR{Parser} which will make this combinator compositional
+(as opposed to \parser{some} defined in Section~\ref{parser:some}). Ultimately
+all of this complexity only shows up in the implementation of our library:
+the end user can happily ignore these details.
 

### 2017-10-27 doc/paper/main.tex
[ doc ] hchainl

@@
 \parser{expr'} ones which are left-associated chains of \parser{base}
-expressions connected by '+'.
+expressions connected by '+'.\label{parser:expr}
 
@@
 
-\subsection{Left and Right Chains}
+\section{Left and Right Chains}
 
@@
 
-We define \parser{schainl} as the fixpoint of either \parser{rest}
-(whose definition we omit here) or the \AR{Success} that was passed
-as an input (the function \AF{\_::$^{r}$} appends a single element
-at the end of a list).
+We define \AF{schainl} as the fixpoint of either \AF{rest} (whose
+definition we omit here) or the \AR{Success} that was passed as an
+input (the function \AF{\_::$^{r}$} appends a single element at the
+end of a list).
 
@@
 the end user can happily ignore these details.
+
+From this definition we can derive \AF{iterate} which takes a parser for
+a seed and a parser for a function and kickstarts a call to \AF{schainl}
+on the result of the parser for the seed.
+
+\agrab{Text/Parser/Combinators}{iterate}
+
+Finally, \AF{hchainl} can be implemented using \AF{iterate}, the applicative
+structure of \AR{Parser} and some of the properties of \BOX{}.
+
+\agrab{Text/Parser/Combinators}{hchainl}
+
+As we have mentioned when defining \AF{schainl}, the combinator \AF{hchainl}
+we have just implemented does not expect fully-defined parsers as arguments.
+As a consequence it can be used inside a fixpoint construction. Both the parser
+for the constructor and the one for its \AB{B} argument are guarded whilst the
+one for the \AB{A} seed is not. This means that trying to define a left-recursive
+grammar by immediately using a recursive substructure on the left is now a type
+error. But it still possible to have some on the right or after having consumed
+at least one character (typically an opening parenthesis, cf. the \type{Expr}
+example in Section~\ref{parser:expr}).
 

### 2017-10-27 doc/paper/abstract.tex
[ admin ] switching to easychair style

--- 2017-10-27 doc/paper/main.tex
@@
-%% For double-blind review submission
-\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission
-%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission
-%\documentclass[acmlarge]{acmart}\settopmatter{}
-
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format should change 'acmlarge' to
-%% 'sigplan,10pt'.
-
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-\usepackage{agda}
-\usepackage{todonotes}
-\usepackage{catchfilebetweentags}
-
-\DeclareUnicodeCharacter{8759}{:\,:}
-\DeclareUnicodeCharacter{954}{$\kappa$}
-
-\newcommand{\shgrab}[1]{\medskip\ExecuteMetaData[parser.tex]{#1}}
-\newcommand{\hgrab}[1]{\ExecuteMetaData[parser.tex]{#1}}
-
-\newcommand{\sagrab}[2]{\medskip\ExecuteMetaData[#1.tex]{#2}}
-\newcommand{\agrab}[2]{\ExecuteMetaData[#1.tex]{#2}}
-\newcommand{\BOX}{\AR{□}}
-
-\input{commands.tex}
-\input{lhs2tex-prelude.tex}
-
-\makeatletter\if@ACM@journal\makeatother
-%% Journal information (used by PACMPL format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmYear{2017}
-\acmMonth{1}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\else\makeatother
-%% Conference information (used by SIGPLAN proceedings format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2017}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\fi
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission
-\setcopyright{none}             %% For review submission
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2017}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-\newcommand{\parser}[1]{\textit{#1}}
-\newcommand{\type}[1]{\texttt{#1}}
-
-\begin{document}
-
-%% Title information
-\title{agdarsec -- Total Parser Combinators}         %% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-%\titlenote{with title note}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-%\subtitle{Subtitle}                     %% \subtitle is optional
-%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
-
-%% Author with single affiliation.
-\author{Guillaume Allais}
-%\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-%  \position{Position1}
-  \department{Digital Security}              %% \department is recommended
-  \institution{Radboud University}            %% \institution is required
-%  \streetaddress{Street1 Address1}
-  \city{Nijmegen}
-%  \state{State1}
-%  \postcode{Post-Code1}
-  \country{The Netherlands}
-}
-\email{gallais@cs.ru.nl}          %% \email is recommended
-
-%% Paper note
-%% The \thanks command may be used to create a "paper note" ---
-%% similar to a title note or an author note, but not explicitly
-%% associated with a particular element.  It will appear immediately
-%% above the permission/copyright statement.
-% \thanks{with paper note}                %% \thanks is optional
-                                        %% can be repeated if necesary
-                                        %% contents suppressed with 'anonymous'
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
 \begin{abstract}
@@
 \end{abstract}
-
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-\begin{CCSXML}
-\end{CCSXML}
-
-% \ccsdesc[500]{Software and its engineering~General programming languages}
-% \ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-\keywords{Parser Combinators, Agda, Total Programming}  %% \keywords is optional
-
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
-\maketitle
-
-\section{Introduction}
-
-Parser combinators have made functional languages such as Haskell
-shine. They are a prime example of the advantages Embedded Domain
-Specific Languages provide the end user. She not only has access
-to a library of powerful and composable abstractions but she is also
-able to rely on the host language's existing tooling and libraries.
-She also gets feedback from the static analyses built in the compiler
-(e.g. type and coverage checking) and can exploit the expressivity
-of the host language to write generic parsers thanks to polymorphism
-and higher order functions.
-
-However she only gets the guarantees the host language is willing
-to give. In non-total programming languages such as Haskell this
-means she will not be prevented from writing parsers which will
-unexpectedly fail on some (or even all!) inputs. Handling a
-left-recursive grammar is perhaps the most iconic pitfall leading
-beginners to their doom: a parser never making any progress.
-
-We start with a primer on parser combinators and follow up with the
-definition of a broken parser which is silently accepted by Haskell.
-We then move on to Agda~\cite{norell2009dependently} and introduce combinators to
-define functions by well-founded recursion. This allows us to define
-a more informative notion of parser and give more precise types to
-the combinators commonly used. We then demonstrate that broken parsers
-such as the one presented earlier are rejected whilst typical example
-can be ported with minimal modifications.
-
-\section{A Primer on Parser Combinators}
-
-\subsection{Parser Type}
-
-Let us start by reminding ourselves what a parser is. Although we
-will eventually move to a more generic type, Fritz Ruehr's rhyme
-gives us the \emph{essence} of parsers:
-
-\begin{verse}
-  A Parser for Things\\
-  is a function from Strings\\
-  to Lists of Pairs\\
-  of Things and Strings!
-\end{verse}
-
-This stanza directly translates to the following Haskell type (we
-use a \textbf{newtype} wrapper to have cleaner error messages):
-
-\shgrab{parser}
-
-It is naturally possible to run such a parser and try to extract
-a value from a successful parse. Opinions may vary on whether a
-run with leftover characters can be considered successful or not.
-We decide against it. This is not crucial as both styles can be
-mutually emulated by either providing an 'end of file' parser
-guaranteeing that only runs with no leftovers are successful or
-by extending a grammar with a dummy case consuming the rest of
-the input string.
-
-\shgrab{parse}
-
-\subsection{(Strongly-Typed) Combinators}
-
-The most basic parser is the one that accepts any character.
-It succeeds as long as the input string is non empty and
-returns one result: the tail of the string together with the
-character it just read.
-
-\shgrab{anychar}
-
-However what makes parsers interesting is that they recognize
-structure. As such, they need to reject invalid inputs. The
-parser only accepting decimal digit is a bare bones example.
-It can be implemented in terms of \parser{guard}, a higher
-order parser checking that the value returned by its argument
-abides by a predicate which can easily be implemented using
-functions from the standard library.
-
-\shgrab{guard}
-\hgrab{digit}
-
-These two definitions are only somewhat satisfactory: the
-resut of the \parser{digit} parser is still \emph{stringly-typed}.
-Instead of using a predicate to decide whether to keep the
-value, we can opt for a validation function of type
-\mbox{\textit{a → Maybe b}} which returns
-a witness whenever the check succeeds. To define this
-alternative \parser{guardM} we can once more rely on code
-already part of the standard library.
-
-In our concrete example of recognizing a digit, we want to
-return an \type{Int} corresponding to it. Once more the
-standard library has just the right function to use together
-with \parser{guardM}.
-
-\shgrab{guard2}
-\hgrab{digit2}
-
-\subsection{Code Reuse: Recognizing Structures}
-
-We have seen how we can already rely on the standard library
-of the host language to seemlessly implement combinators.
-We can leverage even more of the existing codebase by noticing
-that the type constructor \type{Parser} is a \type{Functor},
-an \type{Applicative}~\cite{mcbride2008applicative}, a \type{Monad} and also an \type{Alternative}.
-
-% insert alternative instance
-% interesting use case? replicateM?
-
-\subsection{Expressivity: Higher Order Parsers and Fixpoints}
-
-Our first example of a higher order parser was \parser{guardM}
-which takes as arguments a validation function as well as another
-parser and produces a parser for the type of witnesses returned
-by the validation function.
-
-The two parsers \AB{some} and \AB{many} turn a parser for elements
-into ones for non-empty and potentially empty lists of such elements
-respectively. They concisely showcase the power of mutual recursion,
-higher-order functions and the \type{Functor}, \type{Applicative},
-and \type{Alternative} structure.
-
-\shgrab{some}
-\hgrab{many}
-
-\remark{Non-Commutativity} The disjunction combinator is
-non-commutative. As such the definitions of \textit{some} and
-\textit{many} will try to produce the longest list as possible
-as opposed to a flipped version of \text{many} which would start
-by returning the empty list and slowly offer longer and longer
-matches.
-
-\section{The Issue with Haskell's Parser Types}\label{sec:haskellproblem}
-
-The ability to parse recursive grammars by simply declaring
-them in a recursive manner is however dangerous: unlike type
-errors which are caught by the typechecker and partial covers
-in pattern matchings which are detected by the coverage checker,
-termination is not guaranteed here.
-
-The problem already shows up in the definition of \parser{some}
-which will only make progress if its argument actually uses up
-part of the input string. Otherwise it may loop. However this
-is not the typical hurdle programmers stumble upon: demanding
-a non empty list of nothing at all is after all rather silly.
-The issue manifests itself naturally whenever defining a left
-recursive grammar which leads us to introducing the prototypical
-such example: \type{Expr}, a minimal language of arithmetic
-expressions.
-
-\[
-\texttt{Expr} ~::=~ \texttt{<Int>} ~|~ \texttt{<Expr>} ~\text{`+`}~ \texttt{<Expr>}
-\]
-
-The intuitive solution is to simply reproduce this definition by
-introducing an inductive type for \type{Expr} and then defining
-the parser as an alternative between a literal on one hand and a
-sub-expression, the character '+', and another sub-expression on
-the other.
-
-\shgrab{Expr}
-\hgrab{expr}
-
-However this leads to an infinite loop. Indeed, the second
-alternative performs a recursive call to \parser{expr} even
-though it hasn't consumed any character from the input string.
-
-The typical solution to this problem is to introduce two 'tiers'
-of expressions: the \parser{base} ones which can only be whole
-expressions if we consume an opening parenthesis first and the
-\parser{expr'} ones which are left-associated chains of \parser{base}
-expressions connected by '+'.
-
-\shgrab{base}
-\hgrab{expr2}
-
-This approach can be generalised when defining more complex
-languages by having even more tiers, one for each
-\emph{fixity level}, see for instance~\ref{example:expr}.
-An extended language of arithmetic expressions would for
-instance distinguish the level at which addition and
-substraction live from the one at which multiplication and
-division do.
-
-Our issue with this solution is twofold. Firt, although we did
-eventually managed to build a parser that worked as expected,
-the compiler was unable to warn us and guide us towards this
-correct solution. Additionnally, the blatant partiality of
-some of these definitions means that these combinators and
-these types are wholly unsuitable in a total setting. We could, of
-course use an escape hatch and implement our parsers in Haskell
-but that would both be unsafe and mean we would not be able
-to run them at typechecking time which we may want to do if
-we embed checked examples in our software's documentation,
-or use constant values (e.g. filepaths).
-
-\section{Indexed Sets and Course-of-Values Recursion}
-
-Our implementation of Total Parser Combinators is in Agda,
-a total dependently typed programming language and it will
-rely heavily on indexed sets. But the indices will not be
-playing any interesting role apart from enforcing totality.
-As a consequence, we introduce combinators to build indexed
-sets without having to mention the index explicitly. This
-ought to make the types more readable by focusing on the
-important components and hiding away the artefacts of the
-encoding.
-
-The first kind of combinators corresponds to operations on
-sets which are lifted to indexed sets by silently propagating
-the index. We only show the ones we will use in this paper:
-the pointwise arrow and product types and the constant function.
-
-\sagrab{Relation/Unary/Indexed}{arrow}
-\agrab{Relation/Unary/Indexed}{product}
-\agrab{Relation/Unary/Indexed}{constant}
-
-\remark{Mixfix Operators} In Agda underscores correspond to
-positions in which arguments are to be inserted. It may be a
-bit surprising to see infix notations for functions taking
-three arguments but they are only meant to be partially applied.
-\medskip{}
-
-The second kind of combinator turns an indexed set into a set
-by universally quantifying over the index.
-
-\sagrab{Relation/Unary/Indexed}{forall}
-
-\remark{Implicit Arguments} We use curly braces so that the
-index we use is an \emph{implicit} argument we will never
-have to write: Agda will fill it in for us by unification.
-\medskip{}
-
-We can already see the benefits of using these aliases. For instance
-the fairly compact expression \AF{[} (\AF{κ} \AB{P} \AF{⊗} \AB{Q}) \AF{⟶} \AB{R} \AF{]}
-corresponds to the more verbose type
-\AS{∀} \{\AB{n}\} \AS{→} (\AB{P} \AF{×} \AB{Q} \AB{n}) \AS{→} \AB{R} \AB{n}.
-
-Last but not least, the \BOX{} type constructor takes a \AD{ℕ}-indexed
-set and produces the set of valid recursive calls for a function defined
-by course-of-values recursion: \BOX{} \AB{A} \AB{n} associates to each
-\AB{m} strictly smaller than \AB{n} a value of type \AB{A} \AB{m}.
-This construct, analogous to the later modality showing up in Guarded
-Type Theory~\cite{vezzosi2015guarded}, empowers the user to give precise
-types in a total language to programs commonly written in partial ones
-(see e.g. the definition of \textit{fix} below).
-
-\sagrab{Induction/Nat/Strong}{box}
-
-\remark{Record Wrapper} Instead of defining \BOX{} as a
-function like the other combinators, we wrap the corresponding
-function space in a record type. This prevents normalisation from
-unfolding the combinator too eagerly and makes types more readable
-during interactive development.
-
-\remark{Irrelevance} The argument stating that \AB{m} is strictly
-smaller than \AB{n} is preceded by a dot. In Agda, it means that
-this value is irrelevant and can be erased by the compiler.
-In Coq, we would define the relation \AF{\_<\_} in \type{Prop} to
-achieve the same.
-
-\medskip{}
-
-This \BOX{} type combinator has some useful value combinators.
-The first thing we can notice is the fact that \textbf{\BOX{} is a
-functor}; that is to say that given a natural transformation from
-\AB{A} to \AB{B}, we can define a natural transformation from
-\BOX{} \AB{A} to \BOX{} \AB{B}.
-
-\sagrab{Induction/Nat/Strong}{map}
-
-\remark{Copatterns} The definition of \AF{map} uses
-the \BOX{} field named \ARF{call} on the \emph{left hand side}.
-This is a copattern~\cite{abel2013copatterns}, meaning that we
-explain how the definition is \emph{observed} (via \ARF{call})
-rather than \emph{constructed} (via \AIC{mkBox}).
-
-\medskip{}
-
-Because \AF{\_<\_} is defined in terms of \AD{\_≤\_}, \AF{≤-refl}
-which is the proof that \AD{\_≤\_} is reflexive is also a proof
-that any \AB{n} is strictly smaller than \AN{1} \AF{+} \AB{n}.
-We can use this fact to write the following \AF{extract} function:
-
-\sagrab{Induction/Nat/Strong}{extract}
-
-\remark{Counit} The careful reader will have noticed
-that this is not quite the \textit{extract} we would expect from a
-comonad: for a counit, we would need a natural transformation
-between \BOX{} \AB{A} and \AB{A} i.e. a function of type
-\mbox{\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}}. We will not be able
-to define such a function: \BOX{} \AB{A} \AN{0} is isomorphic to
-the unit type so we would have to generate an \AB{A} \AN{0} out
-of thin air. The types \AB{A} for which \BOX{} has a counit are
-interesting in their own right: they are inhabited at every single
-index as demonstrated by \AF{fix} later on.
-
-\medskip{}
-
-Even though we cannot have a counit, we are still able to define
-a comultiplication thanks to the fact that \AF{\_<\_} is transitive.
-
-\sagrab{Induction/Nat/Strong}{duplicate}
-
-\remark{Identifiers in Agda} Any space-free string which
-is not a reserved keyword is a valid identifier. As a consequence
-we can pick suggestive names such as \AB{m<n} for a proof that $m < n$
-(notice the extra spaces around the infix operator $(<)$).
-
-\medskip{}
-
-Exploring further the structure of the functor \BOX{}, we can observe
-that just like it is not quite a comonad, it is not quite an applicative
-functor. Indeed we can only define \AB{pure}, a natural transformation
-of type \mbox{\AF{[} \AB{A} \AF{⟶} \AR{\BOX{}} \AB{A} \AF{]}}, for the types \AB{A}
-that are downwards closed. Providing the user with \AB{app} is however
-possible:
-
-\agrab{Induction/Nat/Strong}{app}
-
-Finally, we can reach what will serve as the backbone of our parser
-definitions: a safe, total fixpoint combinator. It differs from the
-traditional $Y$ combinator in that all the recursive calls have to be
-guarded.
-
-\sagrab{Induction/Nat/Strong}{fix}
-
-If we were to unfold all the type-level combinators and record wrappers,
-the type of \AF{fix} would correspond exactly to strong induction for
-the natural numbers. Hence its implementation also follows the one of strong
-induction: it is a combination of a call to \AF{extract} and an auxiliary
-definition \AF{fix\BOX{}} of type \mbox{\AF{[} \AR{\BOX{}} \AB{A} \AF{⟶}
-\AB{A} \AF{]} \AS{→} \AF{[} \AR{\BOX} \AB{A} \AF{]}}.
-
-\remark{Generalisation}  A similar \BOX{}
-type constructor can be defined for any induction principle relying
-on an accessibility predicate. Which means that a library's types
-can be cleaned up by using these combinators in any situation where
-one had to give up structural induction for a more powerful alternative.
-
-\section{Parsing, Totally}
-
-As already highlighted in Section~\ref{sec:haskellproblem}, \AB{some}
-and \AB{many} can yield diverging computations if the parser they are
-given as an argument succeeds on the empty string. To avoid any such
-issue, we adopt a radical solution: for a parser's run to be considered
-successful, it must have consumed some of its input.
-
-This can be made formal with the \AR{Success} record type: a \AF{Success}
-of type \AB{A} and size {n} is a value of type \AB{A} together with the
-leftovers of the input string of size strictly smaller than \AB{n}.
-
-\agrab{Text/Parser/Success}{success}
-
-\remark{Implicit Field} Just like the arguments to a function can be
-implicit, so can a record's fields. The user can then leave them out
-when building a value and they will be filled in by unification.
-
-\medskip{}
-
-Coming back to Fritz Ruehr's rhyme, we can define our own \AR{Parser}
-type: a parser for things up to size \AB{n} is a function from strings
-of length \AB{m} less than \AB{n} to lists of \AR{Success}es of size \AB{m}.
-
-\agrab{Text/Parser/Combinators}{parser}
-
-Now that we have a precise definition of \AR{Parser}s, we can start
-building our library of combinators. Our first example \AF{anyChar}
-can be defined by copattern-matching and then case analysis on the
-input string: if it is empty then the list of \AR{Success}es is also
-empty, otherwise it contains exactly one element which corresponds to
-the head of the input string and its tail as leftovers.
-
-\agrab{Text/Parser/Combinators}{anyChar}
-
-Unsurprisingly \AF{guardM} is still a valid higher-order combinator:
-filtering out results which do not agree with a predicate is absolutely
-compatible with the consumption constraint we have drawn. To implement
-\AF{guardM} we can once more reuse existing library functions such as
-\AF{gfilter} which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
-provided a predicate \mbox{\AB{A} \AS{→} \AD{Maybe} \AB{B}}.
-
-\agrab{Text/Parser/Combinators}{guardM}
-
-Demonstrating that \AR{Parser} is a functor goes along the same
-lines: using \AD{List}'s and \AR{Success}'s \AF{map}s. Similarly,
-we can prove that it is an \AF{Alternative}: failing corresponds
-to returning the empty list no matter what whilst disjunction is
-implemented using concatenation.
-
-\agrab{Text/Parser/Combinators}{fmap}\vspace*{-2em}
-\agrab{Text/Parser/Combinators}{fail}\vspace*{-2em}
-\agrab{Text/Parser/Combinators}{disjunction}
-
-So far the types we have ascribed to our combinators are, if we ignore
-the $\mathbb{N}$ indices, exactly the
-same as the ones one would find in any other parsec library. In none
-of the previous combinators do we run a second parser on the leftovers
-of a first one. All we do is either manipulate or combine the results
-of one or more parsers run in parallel, potentially discarding some of
-these results on the way.
-
-However when we run a parser \emph{after} some of the input has already
-been consumed, we could safely perform a \emph{guarded} call. This being
-made explicit would be useful when using \AF{fix} to define a parser for
-a recursive grammar. Luckily \AR{Parser} is, by definition, a
-downwards-closed type. This means that we may use very precise types marking
-all the guarded positions with \BOX{}; if the user doesn't need that extra
-power she can very easily bypass the \BOX{} annotations by using \AF{box}:
-
-\agrab{Text/Parser/Combinators}{box}
-
-The most basic example we can give of such an annotation is probably the
-definition of a conjunction combinator \AF{\_<\&>\_} taking two parsers,
-running them sequentially and returning a pair of their results. The second
-parser is given the type \mbox{\BOX{} \AR{Parser} \AB{B}} instead of
-\mbox{\AR{Parser} \AB{B}} which we would expect to find in other parsec libraries.
-
-\agrab{Text/Parser/Combinators}{conjunction}
-
-We can immediately use all of these newly-defined combinators to give
-a safe, total definition of \AF{some} which takes a parser for \AB{A}
-and returns a parser for non-empty lists of \AB{A}s. It is defined as
-a fixpoint and proceeds as follows: it either combines a head and a
-non-empty tail using \AF{$\_::^{+}\_$} or returns a singleton list.
-
-\agrab{Text/Parser/Combinators}{badsome}
-
-\remark{Inefficiency} Unfortunately this definition is inefficient.
-Indeed, in the base case \mbox{\AF{some} \AB{p}} is going to run the parser
-\AB{p} twice: once in the first branch before realising that \AB{rec}
-fails and once again in the second branch. Compare this definition to
-the Haskell version (after inlining \AB{many}) where \AB{p} is run once
-and then its result is either combined with a list obtained by recursion
-or returned as a singleton:
-
-\shgrab{some2}
-
-This inefficiency can be fixed by using a notion of conjunction
-\AF{\_<\&?>\_} which allows the second parser to fail whilst still
-producing a successful run. The parser \mbox{\AF{some} \AB{p}} can then be
-defined as the conjunction of \AB{p} and a potentially failing
-recursive call. The function \AF{cons} (whose definition we omit here)
-either uses \AF{$\_::^{+}\_$} to put the head and the non-empty
-tail together or returns a singleton list if the recursive call
-failed.
-
-\agrab{Text/Parser/Combinators}{conjunction2}
-\agrab{Text/Parser/Combinators}{goodsome}
-
-\remark{Non-compositionality} The higher-order parser \AF{some}
-expects a \emph{fully} defined parser as an argument. This makes
-it impossible to use it as one of the building blocks of a larger,
-recursive parser. Ideally we would rather have a combinator of
-type \mbox{\AF{[} \AF{Parser} \AB{A} \AF{⟶} \AF{Parser} (\AF{List$^+$} \AB{A}) \AF{]}}.
-This will be adressed in the next subsection.
-\medskip{}
-
-As the previous code snippet shows, \AF{\_<\&?>\_} is defined in
-terms of a more fundamental notion \AF{\_\&?>>=\_} which is a
-combinator analogous to a monad's \textit{bind}. On top of running
-two parsers sequentially (with the second one being chosen based on
-the result obtained by running the first), it allows the second one to fail
-and returns both results.
-
-\agrab{Text/Parser/Combinators}{mbind}
-
-This crucial definition makes it possible to port a lot of the Haskell
-definitions where one would use a parser which does not use any of its
-input. And this is possible without incurring any additional cost as
-the optimised version of \AF{some} showed.
-
-\subsection{Left and Right Chains}
-
-The pattern used in the solution presented in Section~\ref{sec:haskellproblem}
-can be abstracted with the notion of an (heterogeneous) left chain which takes
-a parser for a seed, one for a constructor, and one for and argument. The crucial
-thing is to make sure \emph{not} to use the parser one is currently defining as
-the seed.
-
-\shgrab{hchainl}
-
-We naturally want to include a safe variant of this combinator in our
-library. However this definition relies on the ability to simply use
-\textit{return} in case it's not possible to parse an constructor
-additional constructor and argument and that is something we simply
-don't have access to.
-
-\subsection{Examples}
-
-\todo{expr}\label{example:expr}
-
-%% Acknowledgments
-%\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-%  This material is based upon work supported by the
-%  \grantsponsor{GS100000001}{National Science
-%    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-%  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-%  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-%  conclusions or recommendations expressed in this material are those
-%  of the author and do not necessarily reflect the views of the
-%  National Science Foundation.
-%\end{acks}
-
-%% Bibliography
-\bibliography{main.bib}
-
-%% Appendix
-%\appendix
-%\section{Appendix}
-
-\end{document}

### 2017-10-27 doc/paper/main.tex
[ admin ] switching to easychair style

@@
-%% For double-blind review submission
-\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
-%% For single-blind review submission
-%\documentclass[acmlarge,review]{acmart}\settopmatter{printfolios=true}
-%% For final camera-ready submission
-%\documentclass[acmlarge]{acmart}\settopmatter{}
-
-%% Note: Authors migrating a paper from PACMPL format to traditional
-%% SIGPLAN proceedings format should change 'acmlarge' to
-%% 'sigplan,10pt'.
-
-%% Some recommended packages.
-\usepackage{booktabs}   %% For formal tables:
-                        %% http://ctan.org/pkg/booktabs
-\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
-                        %% http://ctan.org/pkg/subcaption
-\usepackage{agda}
-\usepackage{todonotes}
-\usepackage{catchfilebetweentags}
-
-\DeclareUnicodeCharacter{8759}{:\,:}
-\DeclareUnicodeCharacter{954}{$\kappa$}
-
-\newcommand{\shgrab}[1]{\medskip\ExecuteMetaData[parser.tex]{#1}}
-\newcommand{\hgrab}[1]{\ExecuteMetaData[parser.tex]{#1}}
-
-\newcommand{\sagrab}[2]{\medskip\ExecuteMetaData[#1.tex]{#2}}
-\newcommand{\agrab}[2]{\ExecuteMetaData[#1.tex]{#2}}
-\newcommand{\BOX}{\AR{□}}
-
-\input{commands.tex}
-\input{lhs2tex-prelude.tex}
-
-\makeatletter\if@ACM@journal\makeatother
-%% Journal information (used by PACMPL format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmJournal{PACMPL}
-\acmVolume{1}
-\acmNumber{1}
-\acmArticle{1}
-\acmYear{2017}
-\acmMonth{1}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\else\makeatother
-%% Conference information (used by SIGPLAN proceedings format)
-%% Supplied to authors by publisher for camera-ready submission
-\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
-\acmYear{2017}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
-\fi
-
-%% Copyright information
-%% Supplied to authors (based on authors' rights management selection;
-%% see authors.acm.org) by publisher for camera-ready submission
-\setcopyright{none}             %% For review submission
-%\setcopyright{acmcopyright}
-%\setcopyright{acmlicensed}
-%\setcopyright{rightsretained}
-%\copyrightyear{2017}           %% If different from \acmYear
-
-%% Bibliography style
-\bibliographystyle{ACM-Reference-Format}
-%% Citation style
-%% Note: author/year citations are required for papers published as an
-%% issue of PACMPL.
-\citestyle{acmauthoryear}   %% For author/year citations
-
-\newcommand{\parser}[1]{\textit{#1}}
-\newcommand{\type}[1]{\texttt{#1}}
-
-\begin{document}
-
-%% Title information
-\title{agdarsec -- Total Parser Combinators}         %% [Short Title] is optional;
-                                        %% when present, will be used in
-                                        %% header instead of Full Title.
-%\titlenote{with title note}             %% \titlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-%\subtitle{Subtitle}                     %% \subtitle is optional
-%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
-                                        %% can be repeated if necessary;
-                                        %% contents suppressed with 'anonymous'
-
-%% Author information
-%% Contents and number of authors suppressed with 'anonymous'.
-%% Each author should be introduced by \author, followed by
-%% \authornote (optional), \orcid (optional), \affiliation, and
-%% \email.
-%% An author may have multiple affiliations and/or emails; repeat the
-%% appropriate command.
-%% Many elements are not rendered, but should be provided for metadata
-%% extraction tools.
-
-%% Author with single affiliation.
-\author{Guillaume Allais}
-%\authornote{with author1 note}          %% \authornote is optional;
-                                        %% can be repeated if necessary
-%\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
-\affiliation{
-%  \position{Position1}
-  \department{Digital Security}              %% \department is recommended
-  \institution{Radboud University}            %% \institution is required
-%  \streetaddress{Street1 Address1}
-  \city{Nijmegen}
-%  \state{State1}
-%  \postcode{Post-Code1}
-  \country{The Netherlands}
-}
-\email{gallais@cs.ru.nl}          %% \email is recommended
-
-%% Paper note
-%% The \thanks command may be used to create a "paper note" ---
-%% similar to a title note or an author note, but not explicitly
-%% associated with a particular element.  It will appear immediately
-%% above the permission/copyright statement.
-% \thanks{with paper note}                %% \thanks is optional
-                                        %% can be repeated if necesary
-                                        %% contents suppressed with 'anonymous'
-
-%% Abstract
-%% Note: \begin{abstract}...\end{abstract} environment must come
-%% before \maketitle command
-\begin{abstract}
-Parser combinator libraries represent parsers as functions and,
-using higher-order functions, define a DSL of combinators allowing
-users to quickly put together programs capable of handling complex
-recursive grammars. When moving to total functional languages such
-as Agda, these programs cannot be directly ported: there is nothing
-in the original definitions guaranteeing termination.
-
-In this paper, we will introduce a `guarded' modal operator acting
-on types and show how it allows us to give more precise types to
-existing combinators thus guaranteeing totality. The resulting
-library is available online together with various usage examples
-at \url{https://github.com/gallais/agdarsec}.
-\end{abstract}
-
-%% 2012 ACM Computing Classification System (CSS) concepts
-%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-\begin{CCSXML}
-\end{CCSXML}
-
-% \ccsdesc[500]{Software and its engineering~General programming languages}
-% \ccsdesc[300]{Social and professional topics~History of programming languages}
-%% End of generated code
-
-%% Keywords
-%% comma separated list
-\keywords{Parser Combinators, Agda, Total Programming}  %% \keywords is optional
-
-%% \maketitle
-%% Note: \maketitle command must come after title commands, author
-%% commands, abstract environment, Computing Classification System
-%% environment and commands, and keywords command.
-\maketitle
-
 \section{Introduction}
@@
 \todo{expr}\label{example:expr}
-
-%% Acknowledgments
-%\begin{acks}                            %% acks environment is optional
-                                        %% contents suppressed with 'anonymous'
-  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
-  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
-  %% acknowledge financial support and will be used by metadata
-  %% extraction tools.
-%  This material is based upon work supported by the
-%  \grantsponsor{GS100000001}{National Science
-%    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
-%  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
-%  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
-%  conclusions or recommendations expressed in this material are those
-%  of the author and do not necessarily reflect the views of the
-%  National Science Foundation.
-%\end{acks}
-
-%% Bibliography
-\bibliography{main.bib}
-
-%% Appendix
-%\appendix
-%\section{Appendix}
-
-\end{document}

### 2017-10-30 doc/paper/main.tex
[ doc ] ship it!

@@
 These two definitions are only somewhat satisfactory: the
-resut of the \parser{digit} parser is still \emph{stringly-typed}.
+result of the \parser{digit} parser is still \emph{stringly-typed}.
 Instead of using a predicate to decide whether to keep the
@@
 
-\subsection{Code Reuse: Recognizing Structures}
+\subsection{Expressivity: Structures, Higher Order Parsers and Fixpoints}
+
+%\subsection{Code Reuse: Recognizing Structures}
 
 We have seen how we can already rely on the standard library
-of the host language to seemlessly implement combinators.
+of the host language to seamlessly implement combinators.
 We can leverage even more of the existing codebase by noticing
@@
 
-\subsection{Expressivity: Higher Order Parsers and Fixpoints}
+%\subsection{Expressivity: Higher Order Parsers and Fixpoints}
 
@@
 
-\shgrab{some}
+\medskip{}
+
+\begin{minipage}{0.45\textwidth}
+\hgrab{some}
+\end{minipage}
+\begin{minipage}{0.45\textwidth}
 \hgrab{many}
+\end{minipage}
 
-\remark{Non-Commutativity} The disjunction combinator is
+\remark{Non-Commutative} The disjunction combinator is
 non-commutative. As such the definitions of \textit{some} and
@@
 languages by having even more tiers, one for each
-\emph{fixity level}, see for instance~\ref{example:expr}.
+\emph{fixity level}\label{fixity}, see for instance~\ref{example:expr}.
 An extended language of arithmetic expressions would for
@@
 
-Our issue with this solution is twofold. Firt, although we did
+Our issue with this solution is twofold. First, although we did
 eventually managed to build a parser that worked as expected,
 the compiler was unable to warn us and guide us towards this
-correct solution. Additionnally, the blatant partiality of
+correct solution. Additionally, the blatant partiality of
 some of these definitions means that these combinators and
@@
 \sagrab{Relation/Unary/Indexed}{arrow}
+
+\begin{minipage}{0.55\textwidth}
   \agrab{Relation/Unary/Indexed}{product}
+\end{minipage}
+\begin{minipage}{0.30\textwidth}
 \agrab{Relation/Unary/Indexed}{constant}
+\end{minipage}
 
@@
 
-\agrab{Text/Parser/Combinators}{fmap}\vspace*{-2em}
-\agrab{Text/Parser/Combinators}{fail}\vspace*{-2em}
+\agrab{Text/Parser/Combinators}{fmap}
+\begin{minipage}{0.30\textwidth}
+  \agrab{Text/Parser/Combinators}{fail}
+\end{minipage}
+\begin{minipage}{0.55\textwidth}
   \agrab{Text/Parser/Combinators}{disjunction}
+\end{minipage}
 
@@
 
-\remark{Non-compositionality}\label{parser:some} The higher-order parser \AF{some}
+\remark{Non-Compositional}\label{parser:some} The higher-order parser \AF{some}
 expects a \emph{fully} defined parser as an argument. This makes
@@
 type \mbox{\AF{[} \AF{Parser} \AB{A} \AF{⟶} \AF{Parser} (\AF{List$^+$} \AB{A}) \AF{]}}.
-This will be adressed in the next subsection.
+This will be addressed in the next subsection.
 \medskip{}
@@
 
-\section{Left and Right Chains}
+\section{Left Chains}
 
@@
 
-\subsection{Examples}
+\section{Fully Worked-Out Example}
 
-\todo{expr}\label{example:expr}
+From \AF{hchainl}, one can derive \AF{chainl1} which is not heterogeneous
+and uses the same parser for the seed and the constructors' arguments.
+This combinator together with the idea of fixity mentioned in Section~\ref{fixity}
+is typically used to implement left-recursive grammars. Looking up the
+documentation of the \texttt{parsec} library on hackage~\cite{doc:parsec}
+we can find a fine example: an extension of our early arithmetic language.
+
+\shgrab{example}
+
+One important thing to note here is that in the end we not only get a
+parser for the \type{expr}essions but also each one of the intermediate
+categories \type{term} and \type{factor}. Luckily, our library lets us
+take fixpoints of any sized types we may fancy. As such, we can define
+a sized record of parsers for each one of the syntactic categories:
+
+\agrab{Text/Parser/Examples/Expr}{language}
+
+Here, unlike the Haskell example, we decide to be painfully explicit
+about the syntactic categories we are considering: we mutually define
+three inductive types representing \emph{left-associated} arithmetic
+expressions.
+
+\begin{minipage}{0.33\textwidth}
+  \agrab{Text/Parser/Examples/Expr}{expr}
+\end{minipage}
+\begin{minipage}{0.35\textwidth}
+  \agrab{Text/Parser/Examples/Expr}{term}
+\end{minipage}
+\begin{minipage}{0.20\textwidth}
+  \agrab{Text/Parser/Examples/Expr}{factor}
+\end{minipage}
+
+The definition of the parser itself is then basically the same as the
+Haskell one. Contrary to a somewhat popular belief, working in a
+dependently-typed language does not force us to add any type annotation
+except for the top-level one.
+
+\agrab{Text/Parser/Examples/Expr}{parser}
+
+We can notice four minor changes with respect to the Haskell version. Firstly,
+the intermediate parsers need to be declared before being used which effectively
+reverses the order in which they are spelt out. Secondly, the recursive calls
+are now explicit: in the definition of \type{factor}, \AF{expr} is \AF{map}ped
+under the \BOX{} to project the recursive call to the \AD{Expr} \AR{Parser} out
+of \AR{Language}. Thirdly, we use \AF{hchainl} instead of \AF{chainl1} because
+breaking the grammar into three distinct categories leads us to parsing
+\emph{h}eterogeneous left chains. Fourthly, we have to insert calls to \AF{box}
+to lift \AR{Parser}s into boxed ones whenever the added guarantee that the call
+will be guarded is of no use to us.
+
+\section{More Power: Switching to other Representations}
+
+We have effectively managed to take Haskell's successful approach to defining
+a Domain Specific Language of parser combinators and impose type constraints
+which make it safe to use in a total setting. All of which we have done whilst
+keeping the concision and expressivity of the original libraries. A natural
+next question would be the speed and efficiency of such a library.
+
+Although we have been using a concrete type for \AR{Parser} throughout
+this article, our library actually implements a more general one. It uses
+Agda's instance arguments throughout thus letting the user pick the
+representation they like best.
+
+Firstly, there is nothing special about vectors of characters as an input
+type: any sized input off of which one can peel characters one at a time
+would do. Users may instead use Haskell's \type{Text} packaged together
+with an irrelevant proof that the given text has the right length and a
+binding for \textit{uncons}. This should lead to a more efficient memory
+representation of the text being analysed.
+
+Similarly, there is no reason to force the user to get back a \emph{list}
+of successes: any alternative monad will do. This means in particular that
+a user may for instance instrument a parser with a logging ability to be
+able to return good error messages, have a (re)configurable grammar using
+a \type{Reader} transformer or use a \type{Maybe} type if they want to make
+explicit the fact that their grammar is unambiguous.
+
+\section{Related Work}
+
+This work is based on Hutton and Meijer's influential functional pearl~\citeyear{hutton1998monadic}
+which builds on Walder's insight that exception handling and backtracking can be
+realised using a list of successes~\citeyear{wadler1985replace}. Similar Domain
+Specific Languages have been implemented in various functional languages such as
+Scala~\cite{moors2008parser} or, perhaps more interestingly for us, Rust~\cite{couprie2015nom}
+where the added type-level information about ownership can help implement a
+guaranteed zero-copy parser.
+
+\subsection{Total Parser Combinators}
+
+When it comes to total programming languages, Danielsson's library~\citeyear{danielsson2010total}
+is to our knowledge the only attempt so far at defining a library of total
+parser combinators in a dependently-typed host language. He reifies recursive
+grammars as values of a mixed inductive-coinductive type and tracks at the
+type level whether a sub-grammar accepts the empty word and, as a consequence,
+whether one can meaningfully take its fixpoint.
+
+The reified approach allows him to define a grammar's semantics in terms
+of bags of words and prove sound a variant of Brzozowski derivatives~\citeyear{brzozowski1964derivatives}
+as well as study the equational theory of parsers. The current
+implementation, based on the Brzozowski derivatives, is however of complexity
+at least exponential in the size of the input.
+
+Our approach, although not as successful as Danielsson's when it comes
+to certification, is however more lightweight. Using only strong induction
+on the natural numbers, it is compatible with languages a lot less powerful
+than Agda. Indeed there is no need for good support for mixed induction and
+coinduction in the host language. And although we do rely on enforcing invariants
+at the type-level, one could mimic these in languages with even weaker type
+systems by defining an \emph{abstract} \BOX{} and only providing the user with
+our set of combinators which is guaranteed to be safe.
+
+\subsection{Certified Parsing}
+
+Ambitious projects such as CompCert~\cite{leroy2012compcert} providing
+the user with an ever more certified toolchain tend to bring to light
+the lack of proven-correct options for very practical concerns such as
+parsing. Jourdan, Pottier and Leroy's work~\cite{jourdan2012validating}
+fills that gap by certifying the output of an untrusted parser generator
+for LR(1) grammars. This approach serves a different purpose than ours:
+parser combinators libraries are great for rapid prototyping and small,
+re-configurable parsers for non-critical applications.
+
+Bernardy and Jansson have implemented in Agda a fully-certified generalisation
+of Valiant's algorithm~\cite{bernardy2016certified} by deriving it from
+its specification. This algorithm gives the best asymptotic bounds on
+context-free grammar, that is the \type{Applicative} subset tackled by
+parser combinators.
+
+\section{Conclusion and Future Work}
+
+Starting from the definition of ``parsers for things as functions from
+strings to lists of strings and things'' common in Haskell, we have been
+able to (re)define versatile combinators. However the type system was
+completely unable to root out some badly-behaved programs, namely the
+ones taking the fixpoint of a grammar accepting the empty word or non
+well-founded left recursive grammars. Wanting to use a total programming
+language, this led us to a radical solution: rejecting all the parsers
+accepting the empty word. Luckily, it was still possible to recover a
+notion of ``potentially failing'' sub-parses via a \textit{bind}-like
+combinator as well as defining combinators for left chains. Finally we
+saw that this yielded a perfectly safe and only barely more verbose set
+of total parser combinators.
+
+In the process of describing our library we have introduced a set of
+type-level combinators for manipulating indexed types and defining
+values by strong induction. If we want to provide our users with the
+tools to modularly prove some of the properties of their grammars, we
+need to come up with proof combinators corresponding to the value ones.
+As far as we know this is still an open problem.

### 2017-10-31 doc/paper/main.tex
[ fix ] forgot to mention Char -> Tok

@@
 
-Similarly, there is no reason to force the user to get back a \emph{list}
+Secondly, there is no reason to limit ourselves to \type{Char} as the
+unit of information to be processed. Near all of our combinators are
+fully polymorphic over the kind of tokens they can deal with. To run
+the parser, the user will have to declare an instance of the \AR{Tokenizer}
+class corresponding to their use case. The library provides a trivial
+one for \AD{Char}.
+
+Thirdly, there is no reason to force the user to get back a \emph{list}
 of successes: any alternative monad will do. This means in particular that

### 2017-12-10 doc/paper/main.tex
[ done ] following the reviewers' comments

@@
 shine. They are a prime example of the advantages Embedded Domain
-Specific Languages provide the end user. She not only has access
-to a library of powerful and composable abstractions but she is also
+Specific Languages~\cite{hudak1996building} provide the end user. She not only has access
+to a set of powerful and composable abstractions but she is also
 able to rely on the host language's existing tooling and libraries.
-She also gets feedback from the static analyses built in the compiler
+She can get feedback from the static analyses built in the compiler
 (e.g. type and coverage checking) and can exploit the expressivity
@@
 beginners to their doom: a parser never making any progress.
+Other issues one may want guarantees about range from unambiguity
+to complexity with respect to the input's size.
 
@@
 can be ported with minimal modifications.
+
+\remark{Agda-centric} Although we do use some Agda-specific techniques
+in order to have a codebase as idiomatic as possible, we do not expect
+the reader to be well-versed in them. We insert remarks similar to this
+one throughout the paper to clarify confusing points, and give pointers
+to more in-depth explanations to the interested reader.
+
+This work is however not limited to Agda: it can be ported to other
+dependently-typed languages and we have already done so for Coq~\cite{Coq:manual}
+(\url{https://github.com/gallais/parseque}) and
+Idris~\cite{brady2013idris} (\url{https://github.com/gallais/idris-tparsec}).
 
@@
 
-This stanza directly translates to the following Haskell type (we
-use a \textbf{newtype} wrapper to have cleaner error messages):
+This stanza translates to the following Haskell type. We use a \textbf{newtype}
+wrapper to have cleaner error messages:
 
@@
 It is naturally possible to run such a parser and try to extract
-a value from a successful parse. Opinions may vary on whether a
-run with leftover characters can be considered successful or not.
-We decide against it. This is not crucial as both styles can be
-mutually emulated by either providing an 'end of file' parser
-guaranteeing that only runs with no leftovers are successful or
-by extending a grammar with a dummy case consuming the rest of
-the input string.
+a value from a valid run. Opinions may vary on the exact definition
+of a successful parse: should a run with leftover characters or an
+ambiguous result be accepted? We decide against both in the following
+code snippet but it is not a crucial point.
 
 \shgrab{parse}
+
+Once we are equipped with this type of parsers and a function to run
+them, we can start providing some examples of parsers and parser combinators.
 
@@
 structure. As such, they need to reject invalid inputs. The
-parser only accepting decimal digit is a bare bones example.
+parser only accepting decimal digits is a bare bones example.
 It can be implemented in terms of \parser{guard}, a higher
@@
 value, we can opt for a validation function of type
-\mbox{\textit{a → Maybe b}} which returns
-a witness whenever the check succeeds. To define this
-alternative \parser{guardM} we can once more rely on code
-already part of the standard library.
-
-In our concrete example of recognizing a digit, we want to
-return an \type{Int} corresponding to it. Once more the
-standard library has just the right function to use together
-with \parser{guardM}.
+\mbox{\textit{a → Maybe b}} which returns a witness whenever the
+check succeeds. To define this refined version of \parser{guard}
+called \parser{guardM} we can again rely on the standard library:
 
 \shgrab{guard2}
-\hgrab{digit2}
+
+\begin{itemize}
+  \item \textit{traverse f} of type \mbox{\textit{(String, a) → Maybe (String, b)}}
+    takes apart a pair, applies $f$ to the second component and rebuilds the pair
+    \emph{under} the \textit{Maybe} type constructor,
+  \item \textit{fmap} applies this function to all the elements in the list obtained
+    by running the parser $p$,
+  \item and \textit{catMaybes} of type \mbox{\textit{[Maybe (String, b)] → [(String, b)]}}
+    only retains the values which successfully passed the test.
+\end{itemize}
+
+In our concrete example of recognizing a digit, we want to return the
+corresponding \type{Int}. Once more the standard library has just the
+right function to use together with \parser{guardM}: \textit{readMaybe}
+of (specialised) type \mbox{\textit{String → Maybe Int}}.
+
+\shgrab{digit2}
 
@@
 that the type constructor \type{Parser} is a \type{Functor},
-an \type{Applicative}~\cite{mcbride2008applicative}, a \type{Monad} and also an \type{Alternative}.
+an \type{Applicative}~\cite{mcbride2008applicative}, a \type{Monad}
+and also an \type{Alternative}.
 
-% insert alternative instance
-% interesting use case? replicateM?
+\textbf{Functor} means that given a function of the right type, we
+can alter the values returned by a parser. That is, we have a function
+(which corresponds to the infix combinators (<\$>)):
+
+\shgrab{fmap}
+
+\textbf{Applicative} means two things. First, that given a value of type
+\type{a}, we can define a parser for values of type \type{a}. Second, that
+given a parser for a function and a parser for its argument we can run both
+and apply the function to its argument. That is, we have two functions:
+
+\shgrab{applicative}
+
+\textbf{Monad} means that we are entitled to inspect the result of a
+first parser to decide which one to run next. This brings us beyond the
+realm of context-free grammars. That translates into the existence of
+one function:
+
+\shgrab{monad}
+
+\textbf{Alternative} means that for all type \type{a}, \type{Parser a}
+forms a monoid. It allows us to take the disjunction of various parsers,
+the failure of one leading to the next being used.
+
+\shgrab{alternative}
 
@@
 
-Our first example of a higher order parser was \parser{guardM}
+Our first example of a higher order parser was \parser{guard}
 which takes as arguments a validation function as well as another
@@
 and \type{Alternative} structure.
-
 \medskip{}
@@
 
-\remark{Non-Commutative} The disjunction combinator is
-non-commutative. As such the definitions of \textit{some} and
-\textit{many} will try to produce the longest list as possible
-as opposed to a flipped version of \text{many} which would start
-by returning the empty list and slowly offer longer and longer
-matches.
+\remark{Non-Commutative} The disjunction combinator is non-commutative
+as ultimately we obtain a list (and not a set) of possible results. As
+such the definitions of \textit{some} and \textit{many} will try to
+produce the longest list possible as opposed to a flipped version of
+\text{many} which would start by returning the empty list and slowly
+offer longer and longer matches.
 
@@
 in pattern matchings which are detected by the coverage checker,
-termination is not guaranteed here.
+termination is not guaranteed.
 
@@
 expressions if we consume an opening parenthesis first and the
-\parser{expr'} ones which are left-associated chains of \parser{base}
+\parser{expr} ones which are left-associated chains of \parser{base}
 expressions connected by '+'.\label{parser:expr}
@@
 
+This presentation is still sub-optimal; users would traditionally
+be encouraged to used combinators such as \textit{chainl} to avoid
+this issue. We will only discuss later in Section~\ref{sec:leftchains}.
+
 This approach can be generalised when defining more complex
 languages by having even more tiers, one for each
-\emph{fixity level}\label{fixity}, see for instance~\ref{example:expr}.
+\emph{precedence level}, see for instance Section~\ref{example:expr}.
 An extended language of arithmetic expressions would for
 instance distinguish the level at which addition and
-substraction live from the one at which multiplication and
+subtraction live from the one at which multiplication and
 division do.
@@
 Our issue with this solution is twofold. First, although we did
-eventually managed to build a parser that worked as expected,
+eventually manage to build a parser that worked as expected,
 the compiler was unable to warn us and guide us towards this
@@
 we embed checked examples in our software's documentation,
-or use constant values (e.g. filepaths).
+or use compilation-time configuration via e.g. dependent type
+providers~\cite{christiansen2013dependent}.
 
@@
 the pointwise arrow and product types and the constant function.
-
-\sagrab{Relation/Unary/Indexed}{arrow}
+The second kind of combinator corresponds to universal quantification:
+it turns an indexed set into a set.
 
 \begin{minipage}{0.55\textwidth}
+  \agrab{Relation/Unary/Indexed}{arrow}
   \agrab{Relation/Unary/Indexed}{product}
@@
   \agrab{Relation/Unary/Indexed}{constant}
+  \agrab{Relation/Unary/Indexed}{forall}
 \end{minipage}
@@
 three arguments but they are only meant to be partially applied.
-\medskip{}
-
-The second kind of combinator turns an indexed set into a set
-by universally quantifying over the index.
-
-\sagrab{Relation/Unary/Indexed}{forall}
-
 \remark{Implicit Arguments} We use curly braces so that the
@@
 
-We can already see the benefits of using these aliases. For instance
+We can already see the benefits of these aliases. For instance
 the fairly compact expression \AF{[} (\AF{κ} \AB{P} \AF{⊗} \AB{Q}) \AF{⟶} \AB{R} \AF{]}
@@
 
-Last but not least, the \BOX{} type constructor takes a \AD{ℕ}-indexed
-set and produces the set of valid recursive calls for a function defined
-by course-of-values recursion: \BOX{} \AB{A} \AB{n} associates to each
-\AB{m} strictly smaller than \AB{n} a value of type \AB{A} \AB{m}.
-This construct, analogous to the later modality showing up in Guarded
-Type Theory~\cite{vezzosi2015guarded}, empowers the user to give precise
-types in a total language to programs commonly written in partial ones
-(see e.g. the definition of \textit{fix} below).
+Last but not least, we introduce a type constructor which takes a
+\AD{ℕ}-indexed set and produces the set of valid recursive calls
+for a function defined by course-of-values recursion. By analogy
+to modal logic we call it \BOX{} after the ``necessity'' modality
+whose interpretation in a Kripke semantics is eerily similar to
+our type constructor.
 
@@
 \remark{Record Wrapper} Instead of defining \BOX{} as a
-function like the other combinators, we wrap the corresponding
+function like the other combinators, we wrap the
 function space in a record type. This prevents normalisation from
@@
 achieve the same.
-
 \medskip{}
 
-This \BOX{} type combinator has some useful value combinators.
+This construct can also be understood as analogous to the later modality
+showing up in Guarded Type Theory~\cite{vezzosi2015guarded}. It empowers
+the user to give precise types in a total language to programs commonly
+written in partial ones (see e.g. the definition of \textit{fix} below).
 The first thing we can notice is the fact that \textbf{\BOX{} is a
@@
 rather than \emph{constructed} (via \AIC{mkBox}).
-
 \medskip{}
 
-Because \AF{\_<\_} is defined in terms of \AD{\_≤\_}, \AF{≤-refl}
-which is the proof that \AD{\_≤\_} is reflexive is also a proof
-that any \AB{n} is strictly smaller than \AN{1} \AF{+} \AB{n}.
-We can use this fact to write the following \AF{extract} function:
+Because less than (\AF{\_<\_}) is defined in terms of less than
+or equal (\AD{\_≤\_}), \AF{≤-refl} which is the proof that \AD{\_≤\_}
+is reflexive is also a proof that any \AB{n} is strictly smaller than
+\AN{1} \AF{+} \AB{n}. We can use this fact to write the following
+\AF{extract} function:
 
@@
 issue, we adopt a radical solution: for a parser's run to be considered
-successful, it must have consumed some of its input.
+successful, it must have consumed some of its input. Some nullability
+can be recovered later (see Section~\ref{sec:failure}) when defining
+combinators where one of the sub-parses is allowed to fail.
 
@@
 
-\remark{Implicit Field} Just like the arguments to a function can be
-implicit, so can a record's fields. The user can then leave them out
-when building a value and they will be filled in by unification.
-
+\remark{Implicit Field} Like the arguments to a function can be implicit,
+so can a record's fields. The user can leave them out when building a
+value: they will be filled in by unification.
 \medskip{}
@@
 type: a parser for things up to size \AB{n} is a function from strings
-of length \AB{m} less than \AB{n} to lists of \AR{Success}es of size \AB{m}.
+of length \AB{m} less or equal to \AB{n} to lists of \AR{Success}es of size \AB{m}.
 
 \agrab{Text/Parser/Combinators}{parser}
+
+\subsection{Our First Combinators}
 
@@
 compatible with the consumption constraint we have drawn. To implement
-\AF{guardM} we can once more reuse existing library functions such as
-\AF{gfilter} which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
-provided a predicate \mbox{\AB{A} \AS{→} \AD{Maybe} \AB{B}}.
+\AF{guardM} we can once more reuse existing library functions. Relying
+this time on Agda's standard library rather than Haskell's, the set of
+available function is slightly different. We use for instance \AF{gfilter}
+which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
+provided a predicate \mbox{\AB{A} \AS{→} \AD{Maybe} \AB{B}} and combine
+\AF{sequence} and \AR{Success}'s \AF{map} to obtain a function akin to
+\textit{traverse}.
 
@@
 a safe, total definition of \AF{some} which takes a parser for \AB{A}
-and returns a parser for non-empty lists of \AB{A}s. It is defined as
-a fixpoint and proceeds as follows: it either combines a head and a
-non-empty tail using \AF{$\_::^{+}\_$} or returns a singleton list.
+and returns a parser for \AD{List$^+$} \AB{A}, the type of non-empty
+lists of \AB{A}s. It is defined as a fixpoint and proceeds as follows:
+it either combines a head and a non-empty tail using \mbox{\AF{$\_::^{+}\_$} :
+  \AB{A} $\rightarrow$ \AD{List$^+$} \AB{A} $\rightarrow$
+  \AD{List$^+$} \AB{A}} or returns a singleton list.
 
@@
 
-This inefficiency can be fixed by using a notion of conjunction
-\AF{\_<\&?>\_} which allows the second parser to fail whilst still
-producing a successful run. The parser \mbox{\AF{some} \AB{p}} can then be
-defined as the conjunction of \AB{p} and a potentially failing
-recursive call. The function \AF{cons} (whose definition we omit here)
-either uses \AF{$\_::^{+}\_$} to put the head and the non-empty
-tail together or returns a singleton list if the recursive call
-failed.
+\subsection{Failure is Sometimes an Option}\label{sec:failure}
+
+This inefficiency can be fixed by introducing the notion of a potentially
+failing sub-parse. We use the convention that \AF{?} marks the argument
+of a combinator which is allowed to fail e.g. \AF{<\&?>} is the version
+of the conjunction \AF{<\&>} whose second argument is allowed to fail
+whilst \AF{<?\&>} may let its first argument do so. Which, in terms of
+types, translates to:
 
 \agrab{Text/Parser/Combinators}{conjunction2}
+
+The $\textit{some}~\textit{p} \mathbin{<\!\!|\!\!>} \textit{pure}~ [\,]$ pattern
+used in the definition of \textit{some} can be translated in our total
+setting to a recursive call which is allowed to fail. This leads to the following
+rethought definition of \mbox{\AF{some} \AB{p}}. The inefficiency of the previous
+version has disappeared: \AB{p} is run once and dependening on the success
+or failure of the recursive call it is either added to a non-empty list of
+values or returned as a singleton.
+
 \agrab{Text/Parser/Combinators}{goodsome}
@@
 
-As the previous code snippet shows, \AF{\_<\&?>\_} is defined in
-terms of a more fundamental notion \AF{\_\&?>>=\_} which is a
+The potentially failing conjunction combinator \AF{\_<\&?>\_} can be
+generalised to a more fundamental notion \AF{\_\&?>>=\_} which is a
 combinator analogous to a monad's \textit{bind}. On top of running
 two parsers sequentially (with the second one being chosen based on
-the result obtained by running the first), it allows the second one to fail
-and returns both results.
+the result obtained by running the first), it allows the second one to
+fail and returns both results.
 
@@
 
-This crucial definition makes it possible to port a lot of the Haskell
+These definitions make it possible to port a lot of the Haskell
 definitions where one would use a parser which does not use any of its
-input. And this is possible without incurring any additional cost as
-the optimised version of \AF{some} showed.
+input. Instead of encoding a potentially failing parse using the pattern
+$\textit{p} \mathbin{<\!\!|\!\!>} \textit{pure}~ v$, we can explicitly
+use a combinator acknowledging the authorized failure. And this is possible
+without incurring any additional cost as the optimised version of \AF{some}
+showed.
 
-\section{Left Chains}
+\subsection{Left Chains}\label{sec:leftchains}
 
@@
 library. However this definition relies on the ability to simply use
-\textit{return} in case it's not possible to parse an additional
+\textit{pure} in case it's not possible to parse an additional
 constructor and argument and that is something we simply don't have
-access to. This forces us to find the essence of \parser{rest}, the
-auxiliary definition used in \parser{hchainl}: its first argument is
-not just a value, it is a \AR{Success} upon which it builds until it
-can't anymore and simply returns.
+access to.
 
-We define \AF{schainl} as the fixpoint of either \AF{rest} (whose
-definition we omit here) or the \AR{Success} that was passed as an
-input (the function \AF{\_::$^{r}$} appends a single element at the
-end of a list).
+This forces us to find the essence of \parser{rest}, the auxiliary definition
+used in \parser{hchainl}: its first argument is not just a value, it is a
+\AR{Success} upon which it builds until it can't anymore and simply returns.
+We define \AF{schainl} according to this analysis: it is a bare bones version
+of \textit{hchainl}'s \textit{rest} where \textit{con} and \textit{arg} have
+already been replaced by a single \textit{con} function.
 
@@
 
-The type of \AF{rest} may look intimidating but it has this shape because
-we build a list of \AR{Success}es at the \emph{same} index as the input
-\AR{Success} and \AR{Parser} which will make this combinator compositional
-(as opposed to \parser{some} defined in Section~\ref{parser:some}). Ultimately
-all of this complexity only shows up in the implementation of our library:
-the end user can happily ignore these details.
+A key thing to notice is that we build a list of \AR{Success}es at the
+\emph{same} index as the input \AR{Success} and \AR{Parser} which will make
+this combinator compositional (as opposed to \parser{some} defined in
+Section~\ref{parser:some}). This as a cost in terms of the readability of
+the definition of \AF{schainl}. But ultimately all of this complexity only
+shows up in the implementation of our library: the end user can blissfully
+ignore these details.
 
@@
 one for the \AB{A} seed is not. This means that trying to define a left-recursive
-grammar by immediately using a recursive substructure on the left is now a type
-error. But it still possible to have some on the right or after having consumed
+grammar by immediately using a recursive substructure on the left is now a \emph{type
+error}. But it still possible to have some on the right or after having consumed
 at least one character (typically an opening parenthesis, cf. the \type{Expr}
@@
 
-\section{Fully Worked-Out Example}
+\section{Fully Worked-Out Example}\label{example:expr}
 
@@
 and uses the same parser for the seed and the constructors' arguments.
-This combinator together with the idea of fixity mentioned in Section~\ref{fixity}
+This combinator together with the idea of precedence mentioned in Section~\ref{parser:expr}
 is typically used to implement left-recursive grammars. Looking up the
 documentation of the \texttt{parsec} library on hackage~\cite{doc:parsec}
-we can find a fine example: an extension of our early arithmetic language.
+we can find a fine example: an extension of our early arithmetic language
+(corresponding grammar on the right hand side).
 
+\noindent
+\begin{minipage}{0.45\textwidth}
   \shgrab{example}
+\end{minipage}
+\begin{minipage}{0.55\textwidth}
+  \begin{tabular}{lllll}
+    Expr   & := & Term   & | & Term AddOp Expr \\
+    Term   & := & Factor & | & Factor MulOp Term \\
+    Factor & := & <Int>  & | & '(' Expr ')' \\
+    AddOp  & := & '+' & | & '-' \\
+    MulOp  & := & '*' & | & '/' \\
+  \end{tabular}
+\end{minipage}
 
@@
 
-We can notice four minor changes with respect to the Haskell version. Firstly,
-the intermediate parsers need to be declared before being used which effectively
-reverses the order in which they are spelt out. Secondly, the recursive calls
+Although quite close to the Haskell version, We can notice four minor changes:
+
+Firstly, the intermediate parsers need to be declared before being used which effectively
+reverses the order in which they are spelt out.
+
+Secondly, the recursive calls
 are now explicit: in the definition of \type{factor}, \AF{expr} is \AF{map}ped
 under the \BOX{} to project the recursive call to the \AD{Expr} \AR{Parser} out
-of \AR{Language}. Thirdly, we use \AF{hchainl} instead of \AF{chainl1} because
+of \AR{Language}.
+
+Thirdly, we use \AF{hchainl} instead of \AF{chainl1} because
 breaking the grammar into three distinct categories leads us to parsing
-\emph{h}eterogeneous left chains. Fourthly, we have to insert calls to \AF{box}
+\emph{h}eterogeneous left chains.
+
+Fourthly, we have to insert calls to \AF{box}
 to lift \AR{Parser}s into boxed ones whenever the added guarantee that the call
-will be guarded is of no use to us.
+will be guarded is of no use to us. This last point however does not stand in
+Coq nor Idris which have a mechanism to declare implicit coercions and where the
+typechecker can insert these calls to \AF{box} automatically for us.
 
@@
 fully polymorphic over the kind of tokens they can deal with. To run
-the parser, the user will have to declare an instance of the \AR{Tokenizer}
-class corresponding to their use case. The library provides a trivial
-one for \AD{Char}.
+the parser, the user will have to define an appropriate tokenizer for
+their use case. The library provides a trivial one for \AD{Char}.
 
-Thirdly, there is no reason to force the user to get back a \emph{list}
-of successes: any alternative monad will do. This means in particular that
-a user may for instance instrument a parser with a logging ability to be
-able to return good error messages, have a (re)configurable grammar using
-a \type{Reader} transformer or use a \type{Maybe} type if they want to make
-explicit the fact that their grammar is unambiguous.
+Thirdly, there is no reason to force the user to get back a \AD{List}
+of successes: any Functor which is both a Monad and an Alternative
+will do. This means in particular that a user may for instance instrument
+a parser with a logging ability to be able to return good error messages,
+have a (re)configurable grammar using a \type{Reader} transformer or use
+a \type{Maybe} type if they want to make explicit the fact that their
+grammar is unambiguous.
 
@@
 The reified approach allows him to define a grammar's semantics in terms
-of bags of words and prove sound a variant of Brzozowski derivatives~\citeyear{brzozowski1964derivatives}
+of multisets of words and prove sound a variant of Brzozowski derivatives~\citeyear{brzozowski1964derivatives}
 as well as study the equational theory of parsers. The current
@@
 
-Our approach, although not as successful as Danielsson's when it comes
-to certification, is however more lightweight. Using only strong induction
-on the natural numbers, it is compatible with languages a lot less powerful
-than Agda. Indeed there is no need for good support for mixed induction and
-coinduction in the host language. And although we do rely on enforcing invariants
+Our approach, although not able to tackle certification like Danielsson's,
+is however more lightweight. Using only strong induction on the natural
+numbers, it is compatible with languages a lot less powerful than Agda.
+Indeed there is no need for good support for mixed induction and coinduction
+in the host language. And although we do rely on enforcing invariants
 at the type-level, one could mimic these in languages with even weaker type
